---
title: "Dissertation"
format:
  pdf:
    mermaid:
      theme: neutral
execute:
  echo: false
fig-cap-location: top
bibliography: dissertation.bib
csl: apa.csl
---

```{r}
#| message: false
library(ggplot2)
library(ggmice)
```

<!-- # Outline -->

<!-- ## Eisen onderzoek -->

<!-- Bron: [promotiereglement](https://www.uu.nl/sites/default/files/UBD-Promotieregelement-UU-2025-vastgesteld-20-maart-NL.pdf). -->

<!-- > 1. De promovendus is er verantwoordelijk voor dat het onderzoek dat ten grondslag ligt aan het proefschrift voldoet aan de volgende vereisten: -->

<!-- >   a. de promovendus levert een oorspronkelijke bijdrage aan wetenschappelijk onderzoek die de in Nederland gebruikelijke kwaliteitstoetsing door vakgenoten kan doorstaan; -->
<!--     b. de promovendus heeft aangetoond zelfstandig de wetenschappelijke methoden van  -->
<!-- het vakgebied toe te kunnen passen in de ontwikkeling, interpretatie en toepassing van  -->
<!-- nieuwe kennis; -->
<!--     c. de promovendus heeft kennis genomen van en gewerkt met een substantiële ‘body of  -->
<!-- knowledge’, welke in ieder geval omvat de principes en methoden van de internationale  -->
<!-- wetenschapsbeoefening en de theorievorming, methoden en studies van het  -->
<!-- desbetreffende vakgebied; -->
<!--     d. de promovendus beschikt over het vermogen om een omvangrijk project voor de  -->
<!-- ontwikkeling van nieuwe kennis te ontwerpen en te implementeren; -->
<!--     e. de promovendus is in staat om de kennis en methoden van het desbetreffende  -->
<!-- specialisme en/of vakgebied adequaat over te dragen; -->
<!--     f. de promovendus is in staat om de maatschappelijk verantwoordelijkheid ten aanzien  -->
<!-- van het uitvoeren, toepassen en benutten van het eigen onderzoek te dragen. -->

<!-- > 2. Het onderzoek is verricht in overeenstemming met wettelijke en universitaire voorschriften en gedragsregels -->

<!-- \newpage -->

<!-- ## Inhoud proefschrift -->

<!-- ### Introductie -->

<!-- Thema: computationele evaluatie van imputatiemethodologie. -->

<!-- ### H1: Standardized evaluation -->

<!-- - Inhoud hoofdstuk: review paper (manuscript).  -->

<!-- - Referentie: Oberman, H. I., & Vink, G. (2024). Toward a standardized evaluation of imputation methodology. Biometrical Journal, 66(1), 2200107. https://doi.org/10.1002/bimj.202200107 -->

<!-- - Kwaliteitstoetsing (eis 1a) door peer-review. -->

<!--   - Gepubliceerd, 20x geciteerd (bron: Google Scholar).  -->

<!-- ### H2: Missing the point -->

<!-- - Inhoud hoofdstuk: simulatiestudie (manuscript). -->

<!-- - Referentie: Oberman, H. I., van Buuren, S., & Vink, G. (2021). Missing the Point: Non-Convergence in Iterative Imputation Algorithms. https://arxiv.org/abs/2110.11951 -->

<!-- - Kwaliteitstoetsing (eis 1a) door openbaar maken vroege versie (zie ook [Publish Review Curate](https://www.coalition-s.org/blog/peer-reviewed-preprints-and-the-publish-review-curate-model/) model). -->

<!--   - Pre-print gepubliceerd (4p summary presented at ICML). 11x geciteerd (bron: Google Scholar). -->

<!-- ### H3: R package `ggmice` -->

<!-- - Inhoud hoofdstuk: onderzoekssoftware (verwijzing naar software repository/software vignette). -->

<!-- - Referentie: Oberman, H. I. (2022). `ggmice`: Visualizations for ’mice’ with ’ggplot2’. Zenodo. https://doi.org/10.5281/zenodo.6532702 -->

<!-- - Kwaliteitstoetsing (eis 1a) door open bug reporting (GitHub Issues). -->

<!--   - Code openbaar via GitHub, versie gepubliceerd via Zenodo. 5x geciteerd (bron: Google Scholar). -->

<!--   - GitHub Issues as peer review (10x external Issue, 13x external PR). -->

<!--   - CRAN downloads as impact (852x/month; 21k total). plus conda downloads 100+/month: https://anaconda.org/channels/conda-forge/packages/r-ggmice/overview -->

<!-- ### Discussie -->

<!-- Plaatsing proefschrift in bredere ontwikkelingen wetenschap. -->

<!-- \newpage -->


[TODO: create cover, add boilerplate stuff]

<!-- TODO: add CREDIT and publication packages -->

\newpage

# Acknowledgements

[TODO: write; may be added after uploading dissertation to MyPhD]

\newpage


\tableofcontents

[TODO: fix which sections are included]

\newpage

# Samenvatting

[TODO: write; may be added after uploading dissertation to MyPhD]

\newpage

# Glossary

- complete
- imputed
- missing
- observed
- synthetic

# Introduction

<!-- Problem setting: Missing data and MI/FCS are the central methodological context. -->
<!-- Gap: Current evaluation practices (simulation design, convergence diagnostics, visual diagnostics) are fragmented and insufficient. -->
<!-- Conceptual tool: Introduce and define “computational evaluation” framework for imputation. -->
<!-- Research questions and scope and how each chapter answers part of the overarching question. -->

## Why this dissertation?

[Idea by Hidde: personal relevance to draw the reader in. Not just scientific and societal relevance of research, but also: why me?]

- My motivation is to contribute to trust in science. How? By showing uncertainty in data analysis efforts/inferences/graphs
- Methodology and statistics as a tool for expressing/quantifying uncertainty. 
- While it is also a research discipline in itself, we develop tools that are domain-agnostic.
- I want to contribute methodology that facilitates social and biomedical scientists in drawing valid inferences. 

## What is this dissertation about?

[TODO: add take-home message in 1-2 sentences without jargon]

- Algorithms are everywhere. 
- How can we know when to trust the output of algorithms?
  - Theoretical properties: analytically (often impossible)
  - Empirical performance: via simulation studies ('in silico'; the standard, but not standardized)
  - In practice/applied use: evaluate output, sensitivity analyses, ...? (very ad hoc, not systematic)
- In this dissertation, I investigate how the quality of the output of *imputation* algorithms can be evaluated: algorithms with the purpose of generating plausible values that may replace missing values in incomplete data.  

- verification = "confirmation, through the provision of objective evidence, that specified requirements have been fulfilled" (ISO/IEC/IEEE) ... "process of providing objective evidence that the system, software, or hardware and its associated products conform to requirements (e.g., for correctness, completeness, consistency, and accuracy)" (IEEE) ... "Verification in a life cycle context is a set of activities that compares a product of the life cycle against the required characteristics for that product. ... The system has been built right. ... A system could be verified to meet the stated requirements, yet be unsuitable for operation by the actual users." (IEEE)
- validation = "confirmation, through the provision of objective evidence, that the requirements for a specific intended use or application have been fulfilled" (ISO/IEC/IEEE) ... "Validation in a system life cycle context is the set of activities ensuring and gaining confidence that a system is able to accomplish its intended use, goals, and objectives. The right system has been built. Validation demonstrates that the system can be used by the users for their specific tasks." (IEEE)
- verification asks whether the system has been built right whereas validation asks whether the right system has been built

<!-- - The algorithm central to this dissertation is MICE: multivariate imputation by chained equations. The MICE algorithm is developed to generate 'imputed values': synthetic substitutes to replace missing values in incomplete data, based on observed information.  -->

<!-- - MICE is an iterative algorithm. It produces imputations on a variable-by-variable basis. After a full pass through the data, the algorithm cycles back to the first column to replace the imputed value from the first iteration with an updated entry. -->



Missing data problems are timeless, but technological advances have made treating missingness easier than ever. 
A single line of code can seem to solve your missing data problem nowadays. 
But is that solution really what you expect? 
That is what my dissertation is about.

<!-- generate several completed versions of an incomplete dataset, which allows data analysts to handle the completed datasets as if they were completely observed.  -->

This practice of drawing inference by replacing missing values with synthetic substitutes is called 'multiple imputation'. 
Multiple imputation is a popular and flexible procedure that separates the missing data problem from the analysis of scientific interest. 
Imputation allows data analysts to first solve for the missingness, to then be able to analyze the completed data as if they were completely observed. 

<!-- For data analysts, missing data are a nuisance. Missingness stands in the way of unbiased and confidence-valid statistical inferences. 
Whenever there is just one missing value present in a dataset, extrapolation from the sample to the population may be hindered. 
Firstly, because incomplete data always require some treatment for the missingness. 
Even methods that seemingly ignore the missing data, such as complete case analysis, are a type of missing data treatment. Missing data treatment can severely impact the accuracy and validity of any subsequent analyses. 
Therefore, missing data methods should be applied carefully. 
Secondly, because many missing data methods lead to biased estimates if the incomplete data are in any way different from the complete data one would have liked to obtain. 
For example, when drop-out in a clinical trial is related to the severity of side-effects from the treatment.  -->

<!-- Whenever missing data cannot be prevented, the only way forward is to treat the missing data problem. Missing values force data analysts to treat the missingness in some way.  -->

<!-- Missing data may be widespread, but so are missing data treatment methods.  -->

<!-- Imputation is a convenient and popular missing data method.  -->

Although imputation is widely supported with software, and many implementations of the method have convenient defaults, generating *good* imputations is not a trivial task. 
Evaluating whether these imputations should be trusted remains challenging. 
In this dissertation, I set out to investigate what is needed to develop, apply, and evaluate data-generating algorithms. 
I thereby focus on the imputation procedure 'MICE' as implemented in the R package `mice`.

<!-- - Missing data may be widespread, but so are missing data treatment methods. -->
<!-- - Imputation = filling in missing values to fix the missing data problem. -->
<!-- - The nuisance of missing data is often relieved using imputation methods.  -->
<!-- - To 'impute' means to fill in, so 'imputation' is the practice of replacing missing values with synthetic substitutes.  -->
<!-- - imputation splits the missing data problem and the analysis of scientific interest: first fill in the missingness to be able to analyze the completed data as if they were completely observed -->
<!-- - Imputation is easy with software defaults [e.g. in `mice`, @mice] -->
<!-- - But good imputation is very hard still. -->
<!-- - How do we know if we have a good imputation method? -->
<!-- - In this dissertation, I study computational evaluation of imputation methodology. -->

<!-- Missing data problems are timeless. Technological advances have made imputation fairly easy to apply, but how to apply it *well*? -->

::: {.callout}
TODO: add image for missingness/imputation?
:::

[TODO: add aim of bridging inferential evaluation and data-/algorithm-level diagnostics for imputation methods. current evaluation practices (simulation design, convergence diagnostics, visual diagnostics) are fragmented and insufficient.] $\ell$

<!-- $$Y = \{Y_{obs}, \dot{Y}_{\text{mis}}\}$$ -->

::: {#fig-workflow}

```{mermaid}
%%| fig-width: 6.25
flowchart LR
    subgraph 1["Incomplete data"]
    A(("$$Y$$"))
    end
    subgraph 2["Imputed data"]
    B1("$$\{Y_{\text{obs}}, \dot{Y}_{\text{imp}, \ell}\}$$")
    B2("$$\{Y_{\text{obs}}, \dot{Y}_{\text{imp}, \ell}\}$$")
    B3("$$\{Y_{\text{obs}}, \dot{Y}_{\text{imp}, \ell}\}$$")
    end
    subgraph 3["Analysis results"]
    C1("$$\hat{Q}_{\ell}$$")
    C2("$$\hat{Q}_{\ell}$$")
    C3("$$\hat{Q}_{\ell}$$")
    end
    subgraph 4["Pooled result"]
    D(("$$\bar{Q}$$"))
    end
    
    A --> B1
    A --  impute --> B2
    A --> B3

    B1 --> C1
    B2 -- estimate --> C2
    B3 --> C3

    C1 --> D
    C2 -- pool --> D
    C3 --> D
    
```

*Note*. Incomplete data $Y$ is partially observed, $Y = \{Y_{\text{obs}}, Y_{\text{mis}}\}$. The missing data $Y_{\text{mis}}$ is imputed $m$ times to obtain $m$ completed datasets $\{Y_{\text{obs}}, \dot{Y}_{\text{imp}, \ell}\}$, where $\ell = 1, 2, \dots, m$. On each imputation, the quantity of scientific interest $Q$ is estimated, yielding $m$ analysis results $\hat{Q}_{\ell}$. These results are pooled across imputations to obtain a single pooled result $\bar{Q}$.

MI workflow

:::

<!-- See @fig-workflow. 
%%| fig-cap: "MI workflow"
%%| fig-cap: "MI workflow" -->


```{r}
#| fig-height: 3
load("dat_xy.RData")
# dat_xy[6, 2] <- 5.5
ggmice(dat_xy, aes(X, Y)) + 
  geom_point() + 
  scale_x_continuous(limits = c(1.7, 9), breaks = c(2, 4, 6, 8)) +
  theme_classic() +
  theme(legend.position = "none") +
  annotate(
    geom = "curve", x = 2.5, y = 6.5, xend = 1.8, yend = 5.7, 
    curvature = .3, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate("text", x = 3.6, y = 6.5, label = "missing value")
a <- mice::mice(dat_xy, m = 3, method = "norm", seed = 1, print = FALSE)
ggmice(a, aes(X, Y)) + 
  geom_point() + 
  scale_x_continuous(limits = c(1.7, 9), breaks = c(2, 4, 6, 8)) +
  theme_classic() +
  theme(legend.position = "none") +
  annotate(
    geom = "curve", x = 5, y = 7.5, xend = 4.5, yend = 6, 
    curvature = .3, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate(
    geom = "curve", x = 6, y = 7.4, xend = 5.8, yend = 6, 
    curvature = .2, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate(
    geom = "curve", x = 7.8, y = 7.5, xend = 8.3, yend = 6, 
    curvature = -.3, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate("text", x = 6.5, y = 8, label = "imputed values")  
```

### What is MI?

- MI = replace each missing value with a synthetic substitute multiple times.
- For each missing datapoint, we draw samples from a distribution of values that *might have* been (posterior predictive distribution).
- this generates several imputations: complete*d* datasets.
- completed data can be analyzed as if the data were complete to begin with.
- obtain estimates of the analysis of scientific interest on each imputation.
- estimates are then combined to obtain one pooled estimate [TODO: add SI?].
- the variance of the estimate incorporates not just the sampling variance (within each imputation) but also added variance due to non-response (between imputation variance and simulation error).
- the variability between imputations reflects uncertainty due to the missingness.
- [TODO: why single imputation may yield implausible inference]

::: {.callout}
TODO: add image MI workflow
:::

### What is FCS?

[TODO: reframe as 'Why MI in the flavor of mice?' I.e., FCS is a flexible and agnostic way to specify conditionals to arrive at a joint. MICE also allows JOMO: a direct and inflexible way to arrive at a joint. With MICE, it's easy to generate synthetic values. mice is not just imputation software, but rather an 'information generating machine'. Describe imputation models, analysis models, non-response model and SMC.]

- In general, imputation methods are developed to solve missing data problems.
- What types of missing data problems are we considering in this dissertation? Incomplete samples where some---but not all---entries are missing per row or per column. Entirely missing rows would be 'unit non-response'; entirely missing columns would be unobserved variables. We are interested in data with incomplete cases/variables, or 'item non-response'.
- Missing observations in incomplete cases/variables can be filled in using an imputation model
- In the FCS framework [@vanbuuren2006], we need an imputation model for each incomplete variable 
- fcs is substantive model agnostic, and does not assume any specific joint distribution.
- while "Rubin (Ch. 5) distinguished three tasks for creating imputations under an explicit model: the *modeling* task the *imputation* task and the *estimation* task." [@vanbuuren2007, p. 221], FSC does not require these explicit tasks. Instead, FCS only requires an *imputation model* to describe how synthetic values may be generated, without explicit specification of the joint model.
- imputation models consist of the functional form of the model (e.g., stochastic regression) and imputation model predictors (i.e., other variables in the data)
- recurring questions: how do we choose an appropriate imputation model? and how do we know whether we have chosen an appropriate imputation model?
- "Imputation models bypass the need to specify $P(Y, X, R)$, though their use creates new responsibilities for substantiating its correctness for a given statistical analysis." [@vanbuuren2007, p. 222].
- That's where computational evaluation comes in!

::: {.callout}
# FCS =/= SMC
- the quality of the imputation model can only be determined/assessed based on the complete data model --> TODO: specify terminology!! --> comp eval also possible without analysis model vs statistical evaluation relies on known analysis model (prangender vandaag want imputation gebruikt voor meer analyses) fcs is substantive model agnostic
- what if that model is not (yet) available? then take the most complex possible model and check that, because if you would take the simplest model, then that would implcitly mean you are holding/constraining parameters (forcing b=0), systematically suppressing relations.
- another option would be like smc-fcs: we assume a complete data model and derive the imputations from there.
:::

```{mermaid}
%%| fig-height: 8

flowchart TB
    inc[/"Incomplete data"/] --> miss[\"Inspect missingness"/] 
    miss --> mod[\"`Define imputation model(s)`"/]
    mod --> imp[[Run imputation algorithm]]
    imp --> imps["Imputations<br> $$\quad(1 \dots m)$$"]
    imps --> conv[\"`Inspect convergence`"/]
    conv --> iterate{"`Non-<br>convergence?`"}
    iterate -- "yes" --> imp
    iterate -- "no" --> check[\"`Inspect imputed values`"/]
    check --> misfit{"`Misfit?`"}
    misfit -- "yes" --> mod
    misfit -- "no" --> analyses[\"`Run analyses`"/]
    analyses --> mira["Extract estimates"]
    mira --> pool(["Pool estimates"])
    pool --> mipo["Pooled estimate"]
    inc@{ shape: doc}
    imps@{ shape: docs}
    analyses@{ shape: procs}
    mira@{ shape: tri}
```

::: {.callout}
## TODO: verwerken in flow chart

- misfit en non-conv next to each other? = do the imputation models fit the data?
- Computational evaluation is in every trapezoid: processes that require human intervention. Statistical evaluation is mainly focused on the relation between the pooled estimates and 
- Paragraph after this: convince reviewers that this is the right choice. Giswerk/gebrek aan controle leidt to suboptimale estimates
- missing: draw inference after pooled est; use external info after inspect miss to involva all stakeholders (CRISP-DM step)
- in simulation studies, focus is often just on incomplete data and pooled estimates
:::

### Building and evaluating imputation models

- Model building: how do we know which functional form to choose?
  - rely on defaults, tested in simulation studies 
  - assess the distribution of the incomplete variable (e.g., visual inspection)
  - ...? (maybe ad transformations?)
- Model building: how do we know which imputation model predictors to select?
  - association of incomplete variable with other variables in the data ["include an auxiliary variable if its correlation with the main variable is ≥0.5 (in absolute value) [25]" @nguyen2021]
  - association of missingness indicator with other variables in the data 
  - external input (e.g., branching patterns, or expert knowledge to correct for MNAR)
  - Take the analysis model (if available) as starting point for imputation workflow, for congeneality
  - ...?
- Model evaluation: how do we assess imputation model misfit?
  - fit of the model to the observed data
  - non-convergence in the algorithm 
  - mismatch between observed and imputed data distributions (e.g., visual inspection)
  - ...?

### Evaluating imputations

- FCS has no analytical proof --> rely on evidence from simulation studies
- has been shown that it works under many conditions
- but how to know in practice? we need assumptions
- like assumptions for GLM, we look at the data to evaluate the appropriateness of our chosen model
- some assumption checks are visual inspection, some are quantitative, some need external info
- similarly, we need to evaluate the choice of imputation models

- ideally: evaluate inferential validity, but only available in simulation studies
- in practice, the thing we need is not available by definition
- some assumptions are impossible to definitively determine (e.g. MNAR), some we can evaluate by looking at the data
- this is what this dissertation is about: computational evaluation

- In practice, you cannot validate whether the imputation model was appropriate (???), because what you would need to do so is exactly the thing you don't have: the missing part of the incomplete data, while you only have the observed part.
- The best thing we can do is to rely on a combination of assumptions/theoretical justifications/substantive expertise and indirect evidence of imputation model fit: check the observed versus imputed data distributions, and inspect the imputations for signs of non-convergence. This dissertation offers insight and tools to do so.
- The only setting where we do have the comparative truth (the true but missing values) is simulation studies. 
Imputers have to rely on simulation study results in order to choose the best imputation method for their incomplete variables. 
That's why simulation studies for imputation methodology should be comparable with one another.
Then, imputers can make a 'grounded' assessment which imputation methods may be applicable, and choose the most appropriate one. 
Example: NRI search for lower bound, so impute as 'positive behavior', but per definition wrong statistical inf, because we intentionally bias in one direction

- [in simulation study kun je waarheid maken, maar in de praktijk moet je aannames doen. die aannames zijn belangrijk, juist omdat ze niet/ten dele te bewijzen zijn]
- [what is a simulation study? add a brief methodological background section on simulation studies in statistics]

### What is computational evaluation?

- Computational evaluation is not equal to statistical evaluation. 
Statistical evaluation is important, e.g. to quantify  bias and variance of estimators/procedures, but not the focus of this dissertation.
- Computational evaluation is something to check *conditional on* acceptable statistical properties, and does not replace statistical evaluation. 
- Computational evaluation is originally from information retrieval research [see e.g. @wiki]
- The practice of computational evaluation combines *effectiveness* (i.e., did the process yield the correct result, e.g., accuracy, precision, recall, etc.), *system quality* (e.g., speed, coverage of all possible results, etc.) and, importantly, *user utility* (e.g., happiness, productivity, etc.) [@manning2008]
- The term 'computational evaluation' is not specifically defined for the field of statistics, but it could be useful. 
- One definition of computational evaluation is "to determine the quality of solutions attainable" [@dudek2004] 
- Translated to statistics, a model may be statistically valid, but not fit for purpose. For example, computations that take too long to converge for use in a production pipeline/real-time use in patient care.
- Another definition of computational evaluation is "the process of ensuring an algorithmic solution is a good one: that it is fit for purpose" [@curzon2014].
- We can use this when a statistical process has different purposes and thus different usability/requirements. 
For example, calculating uncertainty at the level of a single case, a sample, or inferring to a population.
- Since the term is not defined for statistics, there is no definition to use for the computational evaluation of imputation methodology.
- In this dissertation, I want to plant a flag: how can we define 'computational evaluation' in the context of imputation methodology?
- Working definition: *systematic assessment of the algorithmic behavior and data-level outputs of imputation procedures*, focused on: algorithmic stability (convergence, failure modes), sensitivity to modeling choices, plausibility and coherence of imputed values, and usability for applied researchers. 
- These aspects are often already taken into account by imputers, but I try to systematize and operationalize these aspects that are often informal or *ad hoc*. 

Dimensions of computational evaluation: e.g. inferential validity, algorithmic behavior, data-level plausibility, user utility, ...

<!-- - which of these is the problem when we use PMM with parse data? on the one hand, it might lead to imputations that are so close that the CI is too narrow and confidence validity is ar risk. on the other hand, it might lead to implausible imputed values for cases with missingness in the sparse region. -->

### Why should we care about computational evaluation of imputation methodology?

- What do we care about? bias/coverage/variance, empirical relevance/plausibility, software-engineering nitpicks like speed/convergence/fault rate.
- Computational evaluation goes beyond statistical evaluation.
- In this dissertation I will use computational evaluation as a suite of tools to investigate the domains of algorithmic stability, data plausibility, and user burden, next to inferential validity.
- Even if we know the process yields correct inferences, we can evaluate the usability/appropriateness/plausibility of the solution. For example, one question to consider is: do the imputed values lie within the range of the observed values?
- Imputation theory does not require the imputed values to be plausible. 
- Rubin [-@rubin1976; -@rubin1987] developed the methodology to get correct population-level inferences from incomplete samples, not at the individual case-level within the sample. [TODO: consider adding that since then, data analysis has changed]
- Bluntly stated: imputation pragmatists/purists don't care for the imputed values themselves, as long as they result in statistically valid inferences.
- But what are the limitations of purely inferential metrics (bias, coverage)? e.g. imputing negative values for age, using PMM for a variable with a sparse region (yielding 'unrealistic' imputed values), or evaluating imputed values of non-converged chain (?), or not knowing the analysis model up front so congeniality is at risk (???) -> PPC would show problem with sparse region
- And vice versa: when might the plausibility of imputed values mislead a data analyst into believing the imputations are statistically valid? when 'looking reasonable' is confused with 'being consistent with the data-generating process', such as single imputation (stoch reg), or imputing under MCAR assumption when there is a MAR mechanism, or destroying the joint distribution of the data while preserving the marginals. [TODO: figure out the difference with Gerko's interpretation of joint occurrence (cross tabs) within conditional distribution; should we make an algorithm to show user which combinations of bivariate obs only occur in imp? Not now!]
- Statistical evaluation of bias and coverage is not possible in practice, only in simulation studies.
- Stat evalualtion may not possible IRL? Bias and confidence interval coverage are impossible to determine without a comparative truth (although PPC and bootstrapping allows us to compare with respect to observed values).
<!-- - statistical evaluation vs computational evaluation: does one come before the other? not necessarily. -->
- [TODO: rephrase as ability to explain differences between observed and imputed data (e.g., lower means in boys data), not as reverse engineering but as statistical evaluation of the data --> then computational eval may be needed to explain differences?] Computational evaluation means not just taking the answer as-is, but tracing back the solution of an algorithm to its origin/back-tracing the procedure: you should *always* be able to retrieve the chain of the imputation procedure. E.g., implausible or impossible imputed values can yield valid statistical inference, but are we satisfied with the way that the imputation were obtained?


<!-- Gerko: "How does computational evaluation differ from statistical eval? IMO it goes beyond it -> e.g. plausible imputations, tracing back the black box to some extend instead of just taking the answer - even if it is correct. You can give examples here wrp nonconverged solutions that yield plausible inference, implausible values that yield valid inference, etcetera."
<!-- Example: {irmi} package by mattias temple yields unbiased estimates, but coverage of 0 because variance was 0 -->
<!-- Gerko: "Would be nice to also consider that taking just a single imputed solution can not be distinguished from another sinle imputed solution, and that the monte carlo variance of the solution should be studied. Therefore it goes way beyond the statistical properties of the imputation methodology." -->
<!-- Hanne:  even if you get the same results based on different SIs, you want to be able to tell which one you should use, that leads us to the need for MI -->
<!-- Gerko: "I think it would be nice to relate it to other stuff/metrics that we value, such as bias/coverage/variance, empirical relevance/plausibility, software-engineering nitpicks like speed/convergence/fault rate." -->

::: {.callout}
# Examples of stat but not comp
- polynomials (ignoring the U in imps could still get stat val)
- example: psych study with baseline, some of which are missing; baseline, intervention, follow-up --> mice would inflate baseline and follow-up if both are incomplete; statistical evaluation will not show this (unbiased, cov ok, etc.); over-imputation or postpred check would show diff distr for obs and imp; circularity should be removed: follow-up should not predict baseline
- fcs iterations not stable, but estimand ok (in simulation study true value) --> conv at cell level??
- PMM with part of sample space with too little donors
:::

### What is the scope of this dissertation?

**RQ: what tools are required to assess the quality of data-generating algorithms? in the development, evaluation, and application.**

- is the status quo sufficient? no, statistical evaluation is required, but not sufficient.
- how Rubin developed the methodology works in theory, but is not enough for contemporary practitioners
- can we use existing tools or do we need more? and how? no, not all are present in the imputer's toolbox. I contribute to the new status quo ("verleggen" = generalizable scientific contribution of this dissertation) 
- unifying how imputations are generated and evaluated (??)
- lifecycle of imputation methods: method is theorized, method is implemented in software, method is evaluated in simulation study, method is put to scrutiny by scientific community, method is applied in practice, method is evaluated in real-world data.
- two types of evaluation: statistical properties of the method in simulation (e.g., bias and confidence-validity; statistical evaluation) & practical application (computational evaluation).

[TODO: remove what I will not address! Maybe add a sentence 'You might expect me to go into...' In the category non-response: just item, and M(C)AR mech. In the category imputation algo: just FCS. In the category substantive models: just GLM for stat inference.] In this dissertation, we will focus on missing data problems with item non-response and MCAR/MAR missingness mechanisms. Unit non-response and MNAR missingness mechanisms are outside of the scope of this dissertation. The missing data method under consideration is primarily multiple imputation by chained equations (MICE), but some results may generalize to other iterative (FCS) imputation algorithms. We do not cover joint modeling (JOMO) imputation or non-imputation missing data methods. Although MICE is not SMC-FCS, we will restrict the analysis of scientific interest mostly to GLM family of models. The applications are aimed at statistical inference (not prediction or causal inference).

## Chapters

1. Evaluation paper: evaluating (new) imputation methods via simulation studies/how we know FCS works 
2. Convergence paper: evaluating imputation algorithm quantitatively
3. Software: choosing/evaluating imputation models in practice


## Introducing chapter 1: evaluation

- **RQ: what to look out for when designing simulation studies for the evaluation of imputation methodology?**/How can computational evaluation criteria be systematically incorporated into simulation studies for imputation methodology?

- published as @oberman2024, cited 27x according to Google Scholar (of which 11x CrossRef)
- based on literature review @liu2023

- How can we trust simulation studies for imputation methods?
- How can we make them compatible & fair?

REF: Morris ADEMP: Using simulation studies to evaluate statistical methods, pre-registration template, FIMD, Introduction to statistical simulations in health research?, Pitfalls and Potentials in Simulation Studies?


## Introducing chapter 2: convergence

- **RQ: FSC is iterative, so convergence *should* be a requirement for valid inference, or is it?**/How does non-convergence in iterative imputation affect inferential validity, and how can non-convergence be diagnosed? 

- simulation study, published as pre-print (11x cited according to Google Scholar), presented at ICML

- even without convergence, we can get valid inference at sample/population level, but not at individual cell level
- imputation stays a custom job: convergence of imputation is conditional on imputation model
- today, imputation are not generated for 1 analysis prob, but all analyses that may be fitted on the data --> all imputation models should be compatible and congenieal --> more 'omvattende' imputation models required
- liefst wil je analysemodel parameters tracken, daarom is computational evaluation zo belangrijk

- Can we quantify the convergence of the MICE algorithm?

REFs: Graphical and numerical diagnostic tools to assess suitability of multiple imputations and imputation models, Diagnostic checking of multiple imputation models, Convergence Properties of a Sequential Regression Multiple Imputation Algorithm, FIMD, Diagnostics for multivariate imputations

## Introducing chapter 3: `ggmice`

[TODO: add poster and 'age old adage' picture vs 1000w.]

This chapter introduces the software `ggmice`: an R package for building and evaluating imputation models. The `ggmice` package offers tools for the visualization of incomplete data, imputation models, and imputed data. 

### Indicators for adoption: how is the software already used?

- software published on Zenodo [@oberman2022], with open code and bug reporting via GitHub
- peer-review not formal via journal, but via Issues and Pull Requests on GitHub (10x external Issue, 13x external PR)
- impact: citations (5x according to Google Scholar), CRAN downloads (852x/month; 21k total) and conda downloads (100+/month; https://anaconda.org/channels/conda-forge/packages/r-ggmice/overview)

### Design principles

- `ggmice` is compatible with the popular R packages `mice` (for imputation) and `ggplot2` (for flexible data visualizations based on the 'grammar of graphics' framework). 
- `mice` imputation models are substantive model agnostic: flexible, var-by-var definition of imputation model. `ggmice` supports this.
- Grammar of Graphics = layered plots [@wilkinson2012] --> `ggmice` produces `ggplot` objects: layered, easily adjustable

- Assumptions: imputation model should fit the observed part, imputation model should be congenial --> providing tools to see observed and imputation side-by-side/evaluation of the distribution of observed and imputed data

### Limitations

- make explicit what these plots/analyses can and cannot tell us: imputation models should fit the observed part of the data, but no way to determinately state miss mech.
- visualization =/= objective truth --> good imputation model fit/plausible values is not a guarantee for statistical validity
- visualization is not a replacement of evaluation but support, also check metrics such as FMI
- interpretation remains subjective (e.g. convergence, MAR assumption, etc.)

### What does ggmice aim to solve?

The `ggmice` package aims to solve the following problems:

 - no direct graphical comparisons between incomplete and imputed data; side-by-side presentation of the data before and after imputation; Visualizations with `ggmice` can be made to inspect marginal and joint distributions of incomplete variables side-to-side to compare incomplete and imputed data
 - no previously existing methods for visualizing `mice` imputation models; complex imputation models are easier to review through visualization (as opposed to console prints/excel exports of predictor matrix)
 - existing plotting tools for `mids` objects not easily editable/publication-ready; Visualizations with `ggmice` can be extended and adapted with tidyverse functions to create publication-ready graphs of imputed data (reflecting the uncertainty due to missingness).
 - unified toolbox for the development and evaluation of imputation models

### Existing solutions

The `ggmice` package fills a hiatus in the software landscape, with other packages generally supporting either `mice` or `ggplot2`, but not both. 
Table REF provides an overview of R packages that offer visualization tools for incomplete and imputed data. In short: with `mice` you cannot plot incomplete data distributions, all other packages lack support for `mice`-imputed data, and visualization of `mice` imputation models did not exist at all prior to `ggmice`.

```{r}
#| echo: false
# library(kableExtra)
packages <- read.csv("pkg_table.csv")
knitr::kable(packages, col.names = c("", names(packages)[-1])) 
```

<!-- https://cranlogs.r-pkg.org/badges/ggmice -->
<!-- https://naniar.njtierney.com/reference/gg_miss_upset.html -->
<!-- https://github.com/njtierney/naniar -->
<!-- https://statistikat.github.io/VIM/articles/VisualImp.html -->
<!-- https://github.com/statistikat/VIM -->
<!-- https://github.com/IQSS/amelia -->

::: {.callout}
## TODO

- write table footnote to explain all aspects of the table. i.e., these are only diffs. the table does not include things that all packages can --> maybe as appendix? 
- explain the purpose of the table: I'm both player and referee. I've got skin in the game, but I made the rubric. Someone else might evaluate/value different aspects.
- also add number of downloads: `ggmice` has 800 monthly compared to `VIM` 13k and `naniar` 22k
- make 'Imputation models' a separate section?
:::


### Software development

- Open Source Software development/FAIR --> R, GitHub, CRAN, Zenodo, ...
- CRAN downloads
- GitHub stars and issues
- StackOverflow mentions/questions
- feedback at conferences
- citations of the package (Zenodo reference)

### Future research & development

- visualizing `mira` objects
- add posterior predictive check plots & over-imputation
- quantifying uncertainty at pattern/cell level
- tools for sensitivity analyses: inducing MNAR in observed part of the data and evaluating effects on the imputations/inference
- What data-level diagnostics are informative for assessing imputation model adequacy in practice (i.e., when inferential validity cannot be evaluated)?
- pre vs post imputation facilitator functies om te checken of er nieuwe verbanden in zitten (kruistabel)


<!-- *Wat is het probleem* <br>  -->

<!-- - Visualization of incomplete data, imputation models, and imputed data.  -->
<!-- - Data exploration for building imputation models: inspect marginal and joint distributions of incomplete variables  -->
<!-- - Evaluation of 'applicablity' imputation models.  -->
<!-- - Visualization of uncertainty due to missingness after imputation.  -->

<!-- *Welke methoden bestaan al* <br> -->

<!-- - `naniar` for incomplete and imputed data (but not `mice`) -->
<!-- - `VIM` for incomplete and imputed data (but not `mice`) -->
<!-- - `mice` for imputed (but not `ggplot2`) -->
<!-- - 'artisinal' code solutions -->
<!-- - print of excel export for imputation models -->

<!-- *Wat zijn de voor- en nadelen van de huidige methoden* <br> -->

<!-- - missing visualization tool compatible with both incomplete data and `mice` imputations: `mice` lacks incomplete data, other packages lack support for `mice`-imputed data -->
<!-- - missing visualization tool for imputation models (i.e., `pred` and `meth` objects) -->

<!-- *Welke nadelen/problemen beoogt de nieuwe aanpak op te lossen* <br> -->

<!-- - no methodology for visualizing `mice` models -->
<!-- - plotting tools for `mids` objects not easily editable/publication-ready -->
<!-- - no direct comparison tool for `mice` between incomplete and imputed data -->

<!-- *In welke opzicht vult de nieuwe methode een lacune* <br> -->

<!-- - unified solution for: missingness, imputation models, imputations -->
<!-- - direct comparison between data before and after imputation with `mice` -->
<!-- - complex imputation models are easier to review through visualization (as opposed to console prints/excel exports of predictor matrix) -->

<!-- *Op welke principes is de nieuwe aanpak gebaseerd* <br> -->

<!-- - Grammar of Graphics: layered plots [@wilkinson2012] -->
<!-- - Open Source Software development/FAIR -->
<!-- - evaluation of the distribution of observed and imputed data -> imputation model should fit the observed part, imputation model should be congenial -->

<!-- *Hoe zijn deze principes in de software vertaald* <br> -->

<!-- - `ggmice` produces `ggplot` objects: layered, easily adjustable -->
<!-- - GitHub, CRAN, Zenodo, ... -->
<!-- - providing tools to see observed and imputation side-by-side -->

<!-- *Welke onderliggende aannames zijn hierbij gebruikt* <br> -->

<!-- - make explicit what these plots/analyses can and cannot tell us: imputation models should fit the observed part of the data, but no way to determinately state miss mech. -->

<!-- *Hoe kun je de huidige en nieuwe gereedschappen het beste met elkaar vergelijken* <br> -->

<!-- - data types (e.g. `mids`) -->
<!-- - how 'publication-ready' the visualizations are (e.g. number of lines of code needed* <br>) -->
<!-- - number of functions* <br>  -->
<!-- - number of downloads* <br> `ggmice` has 800 monthly compared to `VIM` 13k and `naniar` 22k -->

<!-- TODO: table with visualization types in all packages (compare `VIM`, `naniar`, `amelia`) -->


<!-- |                   Functionaliteit                  |  ggmice  |        VIM       |  naniar  |   amelia  |   mice (lattice)   | -->
<!-- |:--------------------------------------------------|:--------:|:----------------:|:--------:|:---------:|:------------------:| -->
<!-- | Visualisatie van de missingness                    | $\checkmark$       | V                | V        | (beperkt) | V                  | -->
<!-- | Visalisatie van het imputatie EN nonresponse model | V        | X                | X        | X         | X                  | -->
<!-- | Imputatie diagnostiek (visueel) + evt in cijfer?   | V        | V (gedeeltelijk) | X        | V         | V                  | -->
<!-- | Ondersteuning mids en andere data typen            | V        | X                | X        | X         | V (alleen lattice) | -->
<!-- | Grammar of Graphics                                | V        | X                | V        | X         | X                  | -->
<!-- | Aanpasbaarheid                                     | hoog     | laag             | hoog     | laag      | laag               | -->
<!-- | Tidyverse compatible?                              | volledig | beperkt          | volledig | beperkt   | geen               | -->
<!-- | Meteen klaar voor publicatie?                      | hoog     | middel           | hoog     | laag      | laag               | -->


<!-- *Levert het nieuwe gereedschap in de praktijk betere resultaten* <br> -->

<!-- - visualization =/= objective truth -> good imputation model fit/plausible values is not a guarantee for statsistical validity -->
<!-- - visualization is not a replacement of evaluation but support, also check metrics such as FMI -->
<!-- - interpretation remains subjective (e.g. convergence, MAR assumption, etc.) -->

<!-- *Wat vinden toekomstige gebruikers van het gereedschap* <br> -->

<!-- - StackOverflow, GitHub issues, feedback at conferences -->

<!-- *Wat zijn relevante indicatoren voor adoptie* <br> -->

<!-- - CRAN downloads -->
<!-- - GitHub stars and issues -->
<!-- - StackOverflow mentions/questions -->
<!-- - feedback at conferences -->
<!-- - citations of the package (Zenodo reference) -->

<!-- *Welke richtlijnen en advies gelden bij gebruik* <br> -->

<!-- - not a replacement of evaluation but support -->

<!-- *Wat zijn beperkingen van de methode* <br> -->

<!-- - interpretation remains subjective (e.g. convergence, MAR assumption, etc.) -->

<!-- *Wat zijn vragen voor nader onderzoek* <br> -->

<!-- - visualizing `mira` objects -->
<!-- - add posterior predictive check plots & over-imputation -->
<!-- - quantifying uncertainty at pattern/cell level -->
<!-- - tools for sensitivity analyses -->
<!-- - What data-level diagnostics are informative for assessing imputation model adequacy in practice (i.e., when inferential validity cannot be evaluated)? -->
<!-- - Do we get better imputations in practice? -->

\newpage

# Chapter 1: evaluation

[TODO: link/insert]

\newpage

# Chapter 2: convergence

[TODO: link/insert]

\newpage

# Chapter 3: ggmice

[TODO: link/insert]

\newpage

# Discussion

<!-- Goal: -->
<!-- consolidate contributions -->
<!-- interpret, not restate—findings -->
<!-- acknowledge limitations -->
<!-- articulate future research agenda -->

## Conclusion

- Computational evaluation of imputation methodology is never finished.

- The main findings in this thesis are...
  - Defining and testing new imputation methods through simulation studies requires careful consideration of the simulation design and evaluation to make imputation methods comparable across studies. 
  We need simulation studies to know imputation methods work. 
  But we cannot be certain they actually *did* work whenever we apply them. The missing part of the data is by definition not available, so we cannot definitively state anything about the imputation model fit to the unobserved part of the data. Moreover, there's untestable assumptions (e.g., MNAR) and lack of clear diagnostic cut offs (e.g., convergence).
  - Non-convergence is hard to diagnose, and typical thresholds to evaluate non-convergence may not be applicable to FCS algorithms: MICE reaches stable output before non-convergence metrics would indicate so. 
  There is a mismatch between theoretical convergence criteria and practical algorithmic behavior. 
  That's why we still have to rely on visual inspection.
  - Visual inspection aids imputers in developing and evaluating imputation models, because formal tests are not available. 
  The R package `ggmice` offers tools for the computational evaluation of the imputations models that were previously unavailable.

- Recommendations for fellow missing data methodologists:
  - design simulation studies structured and reproducible (e.g., report design choices, publish code)
  - talk to your intended audience (i.e., applied researchers), or at least check what they are struggling with (e.g., StackOverflow)
  - ...?

- Recommendations for fellow imputers:
  - think before you code 
  - look at the data, plot the data
  - ...?

## Limitations

- Computational evaluation is not a substitute for thinking. It may give a false sense of certainty, e.g. against MNAR mechanism (untestable assumption).
- Instead, use computational evaluation as complement to (not a replacement for) substantive knowledge and sensitivity analyses. Use expert insight


- This entire dissertation relies on the R language and centers the `mice` package [TODO rephrase as not a flaw: I focus on a dominant ecosystem, and concepts generalize beyond].
  - While this set-up is very popular, the data science landscape has expanded and is ever evolving
  - Many machine learning and deep learning methods are not (yet) implemented as imputation models in `mice`. These might be able to better approximate the posterior predictive distribution of the missing values than `mice` methods, but research should show that still.

- ~~I haven't tested whether the tools I developed actually improve the imputations of `mice` users~~ This dissertation evaluates tools and diagnostics at the methodological level; assessing their causal impact on user behavior and imputation quality is not feasible to systematically test.
  - GitHub issues and StackOverflow are indication
  - Onmogelijk om evidence-based conclusies te trekken, maar misschien wel evidence-informed?

- The visualizations implemented in `ggmice` are generally better suited for numeric data, as opposed to categorical variables (or even open text field data, which is not supported by `mice` currently). Future research on: associations of cat vars with miss indicators, and convergence parameter other than SD of imputed values.

::: {.callout}
- TODO: rephrase `haven't's into future research (positive phrasing: reframe as scope conditions, not shortcomings)
- maybe future: computational evaluation of variable importance within imputation models
:::

## Outlook

- This chapter ends with a future perspective...

- Missing data remains a problem indefinitely. I expect imputation methodology to be developed and applied for the foreseeable future. But the way we deal with data might change.
- With the rise of machine learning and deep learning methods (or "AI"), evaluation will become ever more important. Novel methods are being developed under a prediction framework, not statistical inference framework: ignoring the uncertainty in the estimates, and thus not incorporating between-imputation variance. This leads to too narrow CIs, and too low p-values. Which, in turn, causes spurious results.
- AI models should propagate uncertainty!

- trust in science requires honest reflection of uncertainty in our analyses
- be open about uncertainty, and limitations of scientific studies 
- publish null results (e.g., registered reports or pre-registrations)
- share data and code with other scientists, and the public
- publish open access, educational materials too!

- change the way we evaluate science: this dissertation is an example of the new recognition and rewards vision. it includes software as an actual chapter, not something extra.
- I hope to inspire others to reflect on their own view of what scientific output is. And most of all, I hope to set an example for other PhD candidates to do the same

- imputation is not always the way to go: missingness can tell you about mismatch between data collection method and lived experience --> positionality??

\newpage

# References

::: {#refs}
:::

\newpage

# CV

[TODO: write]
