---
title: "Dissertation"
format:
  pdf:
      include-in-header:
        text: |
          \usepackage[labelfont={bf}]{caption}
          \usepackage{ragged2e}
          \newcommand{\fignote}[1]{\justifying{\footnotesize{\noindent\textit{Note. }{#1}}}}
      mermaid:
        theme: neutral
toc: false
fig-cap-location: top
execute:
  echo: false
bibliography: dissertation.bib
csl: apa.csl
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false
library(ggplot2)
library(ggmice)
load("dat_xy.RData")
```

<!-- `\fignote{...}`{=latex}  -->

<!-- figure captions: https://stackoverflow.com/questions/73775001/quarto-pdf-document-figure-caption-size -->
<!-- # Outline -->

<!-- ## Eisen onderzoek -->

<!-- Bron: [promotiereglement](https://www.uu.nl/sites/default/files/UBD-Promotieregelement-UU-2025-vastgesteld-20-maart-NL.pdf). -->

<!-- > 1. De promovendus is er verantwoordelijk voor dat het onderzoek dat ten grondslag ligt aan het proefschrift voldoet aan de volgende vereisten: -->

<!-- >   a. de promovendus levert een oorspronkelijke bijdrage aan wetenschappelijk onderzoek die de in Nederland gebruikelijke kwaliteitstoetsing door vakgenoten kan doorstaan; -->
<!--     b. de promovendus heeft aangetoond zelfstandig de wetenschappelijke methoden van  -->
<!-- het vakgebied toe te kunnen passen in de ontwikkeling, interpretatie en toepassing van  -->
<!-- nieuwe kennis; -->
<!--     c. de promovendus heeft kennis genomen van en gewerkt met een substantiële ‘body of  -->
<!-- knowledge’, welke in ieder geval omvat de principes en methoden van de internationale  -->
<!-- wetenschapsbeoefening en de theorievorming, methoden en studies van het  -->
<!-- desbetreffende vakgebied; -->
<!--     d. de promovendus beschikt over het vermogen om een omvangrijk project voor de  -->
<!-- ontwikkeling van nieuwe kennis te ontwerpen en te implementeren; -->
<!--     e. de promovendus is in staat om de kennis en methoden van het desbetreffende  -->
<!-- specialisme en/of vakgebied adequaat over te dragen; -->
<!--     f. de promovendus is in staat om de maatschappelijk verantwoordelijkheid ten aanzien  -->
<!-- van het uitvoeren, toepassen en benutten van het eigen onderzoek te dragen. -->

<!-- > 2. Het onderzoek is verricht in overeenstemming met wettelijke en universitaire voorschriften en gedragsregels -->

<!-- \newpage -->

<!-- ## Inhoud proefschrift -->

<!-- ### Introductie -->

<!-- Thema: computationele evaluatie van imputatiemethodologie. -->

<!-- ### H1: Standardized evaluation -->

<!-- - Inhoud hoofdstuk: review paper (manuscript).  -->

<!-- - Referentie: Oberman, H. I., & Vink, G. (2024). Toward a standardized evaluation of imputation methodology. Biometrical Journal, 66(1), 2200107. https://doi.org/10.1002/bimj.202200107 -->

<!-- - Kwaliteitstoetsing (eis 1a) door peer-review. -->

<!--   - Gepubliceerd, 20x geciteerd (bron: Google Scholar).  -->

<!-- ### H2: Missing the point -->

<!-- - Inhoud hoofdstuk: simulatiestudie (manuscript). -->

<!-- - Referentie: Oberman, H. I., van Buuren, S., & Vink, G. (2021). Missing the Point: Non-Convergence in Iterative Imputation Algorithms. https://arxiv.org/abs/2110.11951 -->

<!-- - Kwaliteitstoetsing (eis 1a) door openbaar maken vroege versie (zie ook [Publish Review Curate](https://www.coalition-s.org/blog/peer-reviewed-preprints-and-the-publish-review-curate-model/) model). -->

<!--   - Pre-print gepubliceerd (4p summary presented at ICML). 11x geciteerd (bron: Google Scholar). -->

<!-- ### H3: R package `ggmice` -->

<!-- - Inhoud hoofdstuk: onderzoekssoftware (verwijzing naar software repository/software vignette). -->

<!-- - Referentie: Oberman, H. I. (2022). `ggmice`: Visualizations for ’mice’ with ’ggplot2’. Zenodo. https://doi.org/10.5281/zenodo.6532702 -->

<!-- - Kwaliteitstoetsing (eis 1a) door open bug reporting (GitHub Issues). -->

<!--   - Code openbaar via GitHub, versie gepubliceerd via Zenodo. 5x geciteerd (bron: Google Scholar). -->

<!--   - GitHub Issues as peer review (10x external Issue, 13x external PR). -->

<!--   - CRAN downloads as impact (852x/month; 21k total). plus conda downloads 100+/month: https://anaconda.org/channels/conda-forge/packages/r-ggmice/overview -->

<!-- ### Discussie -->

<!-- Plaatsing proefschrift in bredere ontwikkelingen wetenschap. -->

<!-- \newpage -->


[TODO: create cover, add boilerplate stuff]

<!-- TODO: add CREDIT and publication packages -->

\newpage

# Acknowledgements

[TODO: write; may be added after uploading dissertation to MyPhD]

\newpage

<!-- # TOC -->

```{=latex}
\clearpage
{
\hypersetup{linkcolor=black}
\setcounter{tocdepth}{2}
\tableofcontents
}
```

[TODO: fix which sections are included]

\newpage

# Samenvatting

[TODO: write; may be added after uploading dissertation to MyPhD]

\newpage

# Glossary

[TODO: complete]

- $Y$ = complete data (sized $n \times p$) = $Y=(Y_\mathrm{obs},Y_\mathrm{mis})$ = hypothetically complete data
- $n$ = number of units (indexed $i=1,\dots,n$)
- $p$ = number of variables (indexed $j=1,\dots,p$)

- Incomplete data = $Y_\mathrm{obs}$ = observed data (elements of $y_{ij}$ where $r_{ij}=1$)
- Missing data = $Y_\mathrm{mis}$ = missing data (elements of $y_{ij}$ where $r_{ij}=0$)
- Response indicator (matrix) = $R$ = response indicator (sized $n \times p$) = The matrix $R$ stores the locations of the missing data in $Y.$ 

- Complete-data model = the model of scientific interest/the model we would like to fit if we had complete data
- Missing data model = $P(R|...)$ = missing data model (simplified) = The model that explains what is observed, or 'probability to be missing' (often not of direct scientific interest), such as MCAR ($P(R|\psi)$), MAR ($P(R|Y^\mathrm{obs}, X, \psi)$), and MNAR ($P(R|Y^\mathrm{obs}, Y^\mathrm{mis}, X, \psi)$), where $\psi$ = 	parameters of the missing data model that relates $Y$ to $R$ and $X$ = set of covariates, possibly a subset of $Y$

- MCAR: Missing Completely at Random

- imputed
- observed
- synthetic

\newpage

# Introduction

<!-- Problem setting: Missing data and MI/FCS are the central methodological context. -->
<!-- Gap: Current evaluation practices (simulation design, convergence diagnostics, visual diagnostics) are fragmented and insufficient. -->
<!-- Conceptual tool: Introduce and define “computational evaluation” framework for imputation. -->
<!-- Research questions and scope and how each chapter answers part of the overarching question. -->

<!-- ### TL;DR -->


<!-- [TODO: add take-home message in 1-2 sentences without jargon] -->

### TL;DR (or: the quality of the solution)


<!-- Algorithms are everywhere.  -->

This dissertation is about algorithms and the quality of their output. 
I investigate how a specific type of data-generating algorithm can be evaluated for use in statistical inferences. 
The type algorithm in question are imputation algorithms. 
Imputation algorithms are algorithms with the purpose of generating plausible values based on a model of some observed information. 
This type of algorithm is popular for solving missing data problems.  

<!-- The data-generating algorithms in question are 'imputation' algorithms. -->
<!-- - The algorithm central to this dissertation is MICE: multivariate imputation by chained equations. The MICE algorithm is developed to generate 'imputed values': synthetic substitutes to replace missing values in incomplete data, based on observed information.  -->

<!-- - MICE is an iterative algorithm. It produces imputations on a variable-by-variable basis. After a full pass through the data, the algorithm cycles back to the first column to replace the imputed value from the first iteration with an updated entry. -->


<!-- generate several completed versions of an incomplete dataset, which allows data analysts to handle the completed datasets as if they were completely observed.  -->

Missing data problems are timeless, but technological advances have made treating missingness easier than ever. 
A single line of code can seem to solve your missing data problem nowadays. 
But is that solution really what you expect? 
That is what my dissertation is about.
I want to contribute methodology that facilitates social and biomedical scientists in drawing valid inferences from incomplete data. 

Imputation algorithms are designed to 'impute' (i.e., fill in) missing values in incomplete datasets. 
Incomplete data occur often in scientific disciplines with human subjects, such as the social science and biomedical research.
A participant might, for example, drop out of a longitudinal study, or refuse to answer a sensitive question, resulting in gaps in the research data.
These gaps, or missing values, complicate data analysis and may lead to biased and invalid study results.
Imputation is a popular and flexible procedure that separates the missing data problem from the analysis of scientific interest. 
Imputation allows data analysts to first solve for the missingness, to then be able to analyze the completed data as if they were completely observed. 

The practice of drawing inference by replacing missing values with several synthetic substitutes is called 'multiple imputation'. 
@fig-imp provides an example of multiple imputation, where a missing value is replaced several times with plausible substitutes.
The variability between imputations will reflect the uncertainty due to non-response in the final estimate.

```{r}
#| eval: false
#| include: false
knitr::kable(dplyr::arrange(dat_xy, Y))
```

::: {#fig-imp}

```{r}
#| layout-ncol: 2
#| fig-height: 3
# dat_xy[6, 2] <- 5.5
ggmice(dat_xy, aes(X, Y)) + 
  geom_point() + 
  theme_classic() +
  theme(legend.position = "none") +
  annotate(
    geom = "curve", x = 2.5, y = 7.5, xend = 1.8, yend = 6, 
    curvature = .1, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate("text", x = 3.6, y = 8, label = "missing value")
a <- mice::mice(dat_xy, m = 3, method = "norm", seed = 1, print = FALSE)
ggmice(a, aes(X, Y)) + 
  geom_point() + 
  scale_x_continuous(limits = c(1.7, 9), breaks = c(2, 4, 6, 8)) +
  theme_classic() +
  theme(legend.position = "none") +
  annotate(
    geom = "curve", x = 5, y = 7.5, xend = 4.5, yend = 6, 
    curvature = .3, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate(
    geom = "curve", x = 6, y = 7.4, xend = 5.8, yend = 6, 
    curvature = .2, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate(
    geom = "curve", x = 7.8, y = 7.5, xend = 8.3, yend = 6, 
    curvature = -.3, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate("text", x = 6.5, y = 8, label = "imputed values")  
```

<!-- `\fignote{This is plain text turned as LaTeX caption.}`{=latex} -->

Multiple imputation of a missing value

:::

<!-- For data analysts, missing data are a nuisance. Missingness stands in the way of unbiased and confidence-valid statistical inferences. 
Whenever there is just one missing value present in a dataset, extrapolation from the sample to the population may be hindered. 
Firstly, because incomplete data always require some treatment for the missingness. 
Even methods that seemingly ignore the missing data, such as complete case analysis, are a type of missing data treatment. Missing data treatment can severely impact the accuracy and validity of any subsequent analyses. 
Therefore, missing data methods should be applied carefully. 
Secondly, because many missing data methods lead to biased estimates if the incomplete data are in any way different from the complete data one would have liked to obtain. 
For example, when drop-out in a clinical trial is related to the severity of side-effects from the treatment.  -->

<!-- Whenever missing data cannot be prevented, the only way forward is to treat the missing data problem. Missing values force data analysts to treat the missingness in some way.  -->

<!-- Missing data may be widespread, but so are missing data treatment methods.  -->

<!-- Imputation is a convenient and popular missing data method.  -->

Although imputation is widely supported with software and many implementations of the method have convenient defaults, generating *good* imputations is not a trivial task. 
How can we know when to trust the output of imputation algorithms?
The ideal route of studying the formal statistical properties of the method via analytical derivations is often impossible for imputation algorithms. Instead, the de facto standard way is to estimate their empirical performance via simulation studies, but these are not standardized across studies and therefore hard to compare. 
Moreover, simulation study results might not always translate to applied use. 
And for these real-world applications, there is no systematic format for evaluating the practical performance of imputation algorithms, such as diagnostics for algorithmic convergence. 
Ergo, evaluating whether imputations should be trusted remains challenging. 

  <!-- - Theoretical properties: analytically (often impossible) -->
  <!-- - Empirical performance: via simulation studies ('in silico'; the standard, but not standardized) -->
  <!-- - In practice/applied use: evaluate output, sensitivity analyses, ...? (very ad hoc, not systematic) -->
  
In this dissertation, I set out to investigate what is needed to develop, apply, and evaluate data-generating algorithms. 
I thereby focus on the imputation procedure 'MICE' as implemented in the R package `mice`.

<!-- - Missing data may be widespread, but so are missing data treatment methods. -->
<!-- - Imputation = filling in missing values to fix the missing data problem. -->
<!-- - The nuisance of missing data is often relieved using imputation methods.  -->
<!-- - To 'impute' means to fill in, so 'imputation' is the practice of replacing missing values with synthetic substitutes.  -->
<!-- - imputation splits the missing data problem and the analysis of scientific interest: first fill in the missingness to be able to analyze the completed data as if they were completely observed -->
<!-- - Imputation is easy with software defaults [e.g. in `mice`, @mice] -->
<!-- - But good imputation is very hard still. -->
<!-- - How do we know if we have a good imputation method? -->
<!-- - In this dissertation, I study computational evaluation of imputation methodology. -->

<!-- Missing data problems are timeless. Technological advances have made imputation fairly easy to apply, but how to apply it *well*? -->

Why? Because current evaluation practices (simulation design, convergence diagnostics, visual diagnostics) are fragmented and insufficient. 

How? By bridging inferential evaluation and data-/algorithm-level diagnostics for imputation methods.

<!-- $$Y = \{Y_{obs}, \dot{Y}_{\text{mis}}\}$$ -->


### How did we get here? (or: statistical inference is just a tool)

<!-- [Idea by Hidde: personal relevance to draw the reader in. Not just scientific and societal relevance of research, but also: why me?] -->

My academic tag line has long been "statistician, interdisciplinarian, open scientist". 
And while there's no doubt I am an interdisciplinarian and an open scientist, I am no longer keen on calling myself a statistician necessarily.
I am more inclined to say I am a methodologist rather than a statistician. 
As a methodologist, I am interested in how science is done, and how it can be done better.
Especially in the social and biomedical sciences, research should not be purely quantitative. 
Statistics is just *one* way of doing science. 

Don't get me wrong, statistics is still a favorite of mine. I love stats (I even have a mug that says so). 
Statistics is the science of uncertainty and extracting information from data (Hand, 2010).
Extracting information from data and turn them into insights can feel magical. 
Statistical inference allows us to distinguish between real-world effects and random noise (Rosenberg & Curtain, 2015). 
I am most fan of the uncertainty of it all.
<!-- Since statistical inference employs mathematical probability theory (i.e., general laws) to generalize from a sample to a population (i.e., extrapolation), it may be classified as either deductive or inductive inference (respectively).   -->
<!-- There is no consensus in the literature about which pattern of scientific reasoning suits statistical inference (Ruiter & Albert, 2017).  -->
<!-- Gelman & Shalizi argue that statistical inference is deductive in the context of hypothesis testing–when general laws from mathematical probability theory are used–and inductive in the context of parameter estimation–when sample statistics are extrapolated to the population.  -->
<!-- "Statistics is both the science of uncertainty and the technology of extracting information from data." - featured in the International Encyclopedia of Statistical Science (Hand, 2010).  -->
I am motivated to contribute to trust in science by making it easier to reflect uncertainty in data analysis efforts, inferences, and graphs. 
In this dissertation, I hope to achieve this for missing data methodology.

<!-- The topic of this dissertation is missing data. -->
<!-- - Methodology and statistics as a tool for expressing/quantifying uncertainty.  -->
<!-- While statistics is also a research discipline in itself, we develop tools that are domain-agnostic. -->



<!-- See @fig-workflow. 
%%| fig-cap: "MI workflow"
%%| fig-cap: "MI workflow" -->

### Why are missing data problematic?

<!-- We will assume that you took the necessary measures to prevent and/or mend any missing observations during the data collection stage of your research project, but still ended up with an incomplete dataset.  -->
<!-- We will investigate which strategies you may employ when faced with such a missing data problem. -->

<!-- According to Gelman & Shalizi “statistical models are tools that let us draw inductive inferences on a deductive background” (2013, p. 27).  -->

Statistical analyses are a means to infer conclusions about a population after observing only a subset of this population (a sample). 
If the sample is representative, sample estimates will reflect the population characteristics plus some uncertainty surrounding it.
Statisticians call this uncertainty the 'sampling error' due to observing only a subset of the population, which is often expressed as part of the 'standard error' (SE) surrounding the estimates.
Standard errors then get translated into decision tools for inference such as $p$-values and confidence intervals (CIs), which help researchers distinguish between signal and noise. 
So, statistics is a means of estimating effects within a sample and inferring back to the population with some expression of uncertainty. 

Generally speaking, statistical inference works really well. 
For example, when sociologists wanted to study trust in the government during the COVID-19 pandemic, it was not feasible to survey all Dutch citizens. 
Instead, they used a representative sample. 
Some thousands of people were asked to fill in a questionnaire about their trust in the government over the course of the pandemic. 
By repeating the questionnaire over time, the researchers wanted to gain insight into the development of (dis)trust among different subgroups in Dutch society.
The estimates from this sample would have been used to infer back to the population and inform policy, if it had not been for missing data...

Missing data are a nuisance because it is impossible to do statistical inferences with incomplete data. 
When there is just one missing value in a dataset, statistical inference cannot be performed. 
With incomplete data, even simple statistics such as the mean of variables are mathematically undefined.
<!-- It is not possible to use an incomplete sample to infer about a larger population without .  -->
To obtain results from your incomplete dataset, you will always employ a missing data strategy-–whether deliberately or not. 
<!-- Missing data are not a problem with purely descriptive statistics; the problems arise when we want to infer from incomplete sample to the population (or the complete sample).  -->

::: {#fig-mis}

```{mermaid}
%%| fig-height: 2.2

flowchart TD
  A1[population] --sampling--> B1([complete<br> sample])
  B1 --inference--> A1

x:::hidden ~~~ y:::hidden
classDef hidden display: none;

  A2[population] --sampling--> B2([complete<br> sample]) --non-response--> C2([incomplete<br> sample])
  style B2 stroke:#f66,stroke-width:1px,stroke-dasharray: 4
  C2 --inference +<br>missing data<br>method--> A2
```

`\fignote{Schematic view of statistical inference in the presence of non-response. Left: no missing data in the sample from the population (i.e., ideal circumstances/business as usual). Right: incomplete sample due to non-response, so no complete sample could be obtained}`{=latex} 

<!-- ::: {style="font-size: 50%;"} -->



<!-- ::: -->

Inference with incomplete data

:::

<!-- %%| fig-cap: test -->
To get an estimate from incomplete data, some kind of missing data method has to be employed. 
Even without specifying a missing data method explicitly, there will be  handling of the missing data. 
<!-- It's only possible to do descriptive statistics like Statistics Netherlands does, but it is not possible to use a sample to infer about a larger population.  -->
<!-- For example, when sociologists wanted to use a sample to estimate the trust in the government during the Covid pandemic, if there is just one missing data point, it is impossible to infer back to the population.  -->
<!-- To obtain results from your incomplete dataset, you will always employ a missing data strategy-–whether deliberately or not.  -->
The default strategy in most statistical software packages is to ignore all cases that have missing values (Van Buuren, 2018). 
This is referred to as ‘list-wise deletion’ (LWD). 
Ignoring missingness, however, lowers statistical power and may yield invalid inferences (e.g., biased estimates, spurious significant results; Van Buuren, 2018). 
List-wise deletion is ill-advised under most conditions (see van Buuren 2018 for exceptions).
It is better to treat the missing data than to ignore it.

List-wise deletion (complete‑case analysis) assumes that the observed part of the data is representative of the missing part of the data.
List-wise deletion results in bias because it assumes that the observed part of the data is equivalent to the missing part of the data, and this is not usually the case. 
<!-- This corresponds to a missing completely at random (MCAR) mechanism; when MCAR does not hold, listwise deletion yields biased estimates. -->

### Why are the missing and observed parts not equivalent? (or: missing data mechanisms)

[TODO: edit!] In that study about trusting the government during the COVID-19 pandemic, it was not random who had missing data on the question about trust in the government. 
The technical term is 'Missing Not At Random'. 
The missingness depends on unobserved values themselves (i.e., people who distrust the government are less likely to respond to the trust item); this cannot be resolved from the observed data alone and requires assumptions or external information.
If we can model the missingness probability based on known data, for example, previous observations or demographic data, we have 'Missing At Random'.
<!-- We can model the missingness probability using known variables and obtain valid inference if this model is adequate -->
That is, conditional on this information, we can get to 'missing at random' or 'seen data dependent'. 
This is in contrast to the 'unseen data dependent' missingness mechanism or 'missing not at random'. 
If there would be purely random missingness independent of individual characteristics, there is missing completely at random missingness mechanism, for example, due to a glitch in the service software. 
In human subjects research, pure MCAR is unlikely; methods that implicitly assume MCAR (like listwise deletion) are therefore likely to cause bias.
In practice, it is impossible to know from the observed part of the data alone, so we have to make assumptions. 
So don't just use list-wise deletion. 
Treat the missingness to accommodate differences between the observed part of the data and the missing part. 

::: {#fig-mech}
```{mermaid}
%%| fig-width: 6.25
graph TB
linkStyle default interpolate basis
 subgraph MNAR[MNAR]
    X3["X"]
    Y3["Y"]
    Z3["Ψ"]
    R3["R"]

    X3 <--> Y3
    Z3 --> R3
    X3 --> R3
    Y3 --> R3
 end

 subgraph MAR
    X2["X"]
    Y2["Y"]
    Z2["Ψ"]
    R2["R"]

    X2 <--> Y2
    Z2 --> R2
    X2 --> R2
 end

subgraph MCAR
    X1["X"]
    Y1["Y"]
    Z1["Ψ"]
    R1["R"]

    X1 <--> Y1
    Z1 --> R1

 end

```

`\fignote{Y = incomplete data, X = fully observed covariate, R = missingness indicator for Y, $\psi$ = causes of the missingness unrelated to X and Y}`{=latex} 

Missing data mechanisms
:::

### How to treat the missing data problem?

There are three broad categories: deletion based methods (e.g. list-wise deletion), methods that incorporate the missingness into the model of scientific interest (i.e., the analysis model, such as a regression model), and imputation methods.
Deletion-based methods do not generally accomodate differences between the observed and missing parts of the data, making them mostly applicable to MCAR. 
Modeling methods do accommodate MAR, but are often quite complex. 
For example, using 'Bayesian' methods that incorporate the missingness into the analysis model: they jointly specify the data model and missingness model per analysis.
You have to specify the model for each analysis separately, and there are no null hypothesis significance tests which is a problem when working with applied researchers. 
Another option to incorporate the missingness in the analysis model is expectation maximization. 
But this yields many parameters and might not be available for a complex multi-level model like the Covid 19 data. 
Both of these techniques are tied to a specific analysis model, which might not be suitable for contemporary practitioners who run several models in a data science workflow. 
Imputation methods do not have these drawbacks of being computationally advanced and analysis-model specific.

Imputation separates the missing data problem from the analysis model and is the focus of this dissertation.
Why impute?
Imputation splits the missing data problem from the analysis of scientific interest: first solve missingness, then analyze completed data “as if” they were fully observed.
This helps when different people are responsible for different parts of a project (e.g., a missing data specialist vs. domain analyst) and when many different analyses are planned on the same data.
It allows the use of familiar complete‑data methods instead of bespoke incomplete‑data models, and it is widely supported in software: one line of code can resolve the missing data problem.
Compared to listwise deletion, imputation methods can accommodate MAR mechanisms, and reduce research waste by retaining incomplete cases. 
Compared to missing data methods that incorporate the missing data like Bayesian methods or EM-based approaches, imputation is more flexible and often less complex for non‑experts.

### How does imputation work? 

We generate imputations to fill in the missing part of the data. We explicitly model the distribution of this missing part based on what is observed. So we use the observed part of the data to fix the missing part of the data. 
Which types of model or models do we use? 
The simplest model is no model at all.
With hot deck imputation, we use random other person's observation to impute or fill in the values for people with a missing data point. The problem is that this disrupts the multivariate distribution of the data, so we get biased estimates. 

Another simple technique is a univariate statistic such as the mean. Unfortunately, the mean will bias your scientific estimates even worse than list-wise deletion would, because there is no correction for the missing at random missingness mechanism. 
Mean imputation (and other univariate imputation methods) severely biases estimates and fails to account for MAR mechanisms.

<!-- Unfortunately, popular ad-hoc missing data methods, such as replacing missing values with variable means, lead to biased estimates.  -->
Even 'advanced' techniques for substituting missing values that recover the unobserved value almost perfectly are problematic.
Multivariate models such as stochastic regression imputation correct for MAR missingness, but underestimate variance if used as single imputation.
All techniques that rely on a single replacement value risk under-representing uncertainty in the inferences, because the uncertainty due to the non-response is not taken into account.
Thus, repeat stochastic draws to create a distribution of plausible values per missing datapoint, reflecting the uncertainty due to non‑response.
<!-- We will assume that you took the necessary measures to prevent and/or mend any missing observations during the data collection stage of your research project, but still ended up with an incomplete dataset. -->

### Multiple imputation

MI is the practice of replacing each missing value with a synthetic substitute, multiple times.
For each missing datapoint, we draw samples from a distribution of values that *might have* been (posterior predictive distribution).
This generates several imputations: complete*d* datasets.
The completed datasets can be analyzed as if the data were complete to begin with.
On each completed dataset, we obtain estimates of the analysis of scientific interest.
These estimates per imputation are then combined to obtain one pooled estimate.
The variance of the estimate incorporates not just the sampling variance (within each imputation) but also added variance due to non-response (between imputation variance and simulation error).
The variability between imputations reflects uncertainty due to the missingness.

::: {#fig-scheme}

```{mermaid}
%%| fig-width: 6.25
flowchart LR
    subgraph 1["Incomplete data"]
    A(("$$Y$$"))
    end
    subgraph 2["Imputed data"]
    B1("$$\{Y_{\text{obs}}, \dot{Y}_{\text{imp}, \ell}\}$$")
    B2("$$\{Y_{\text{obs}}, \dot{Y}_{\text{imp}, \ell}\}$$")
    B3("$$\{Y_{\text{obs}}, \dot{Y}_{\text{imp}, \ell}\}$$")
    end
    subgraph 3["Analysis results"]
    C1("$$\hat{Q}_{\ell}$$")
    C2("$$\hat{Q}_{\ell}$$")
    C3("$$\hat{Q}_{\ell}$$")
    end
    subgraph 4["Pooled result"]
    D(("$$\bar{Q}$$"))
    end
    
    A --> B1
    A -- " impute" --> B2
    A --> B3

    B1 --> C1
    B2 -- estimate --> C2
    B3 --> C3

    C1 --> D
    C2 -- pool --> D
    C3 --> D
    
```

`\fignote{Read left to right. Incomplete data $Y$ is partially observed, $Y = \{Y_{\text{obs}}, Y_{\text{mis}}\}$. The missing data $Y_{\text{mis}}$ is imputed $m$ times to obtain $m$ completed datasets $\{Y_{\text{obs}}, \dot{Y}_{\text{imp}, \ell}\}$, where $\ell = 1, 2, \dots, m$. On each imputation, the quantity of scientific interest $Q$ is estimated, yielding $m$ analysis results $\hat{Q}_{\ell}$. These results are pooled across imputations to obtain a single pooled result $\bar{Q}$.}`{=latex} 

Multiple imputation scheme ($m=3$)

:::
Multiple imputation means that each missing value is filled in several times, with slightly different model parameters to represent model uncertainty. 
The result of MI is that the uncertainty in the data due to missingness is preserved in the inference. 

<!-- In the context of hypothesis testing, we deduce probabilities (or probability distributions, if you prefer Bayesian statistics) by comparing a hypothesized test statistic to the data that was obtained. The resulting probability is a mathematical property, and thus certain knowledge (Gelman & Shalizi, 2013).  -->
<!-- The interpretation of confidence intervals (CIs), for instance, is the result of a deductive pattern of reasoning.  -->
<!-- It is a mathematical fact that at least 95% of all possible sample CIs from a certain population will contain the true population parameter.  -->
<!-- We do not know, however, whether the CI that we observed in our sample is one of the 95% (or more) that does indeed cover the true population parameter, or that we obtained one of the (less than) 5% CIs without this value.  -->
<!-- And if we don’t know whether the specific CI observed in our sample contains the true population parameter value or not, we cannot deduce the probability that the true population parameter lies within the bounds of the CI that we observed.  -->
<!-- To that end, we would need to extrapolate outside of the observations that we have, i.e., use induction.  -->
<!-- Ergo, if we move from the context of hypothesis testing to the context of estimation, we need induction as our pattern of scientific reasoning. -->

MI enables researchers with missing data problems to correct for differences due to missingness. 
Correcting for missingness happens at two levels: 1) modeling the relation between missing cases and covariates; 2) adding the appropriate variance to the models to represent the uncertainty due to missingness. 

The result is a probability distribution for each missing value, from which we subsequently draw several values as imputations.
If we would use infinitely many values as imputations, we would obtain the entire probability distribution. 
In practice, however, we use a limited number of imputations. 
With that, we approximate the probability distributions of missing values. 
The more imputations are used, the better we can represent the uncertainty in the data due to missingness.


<!-- With inductive inference alone, we can never obtain complete certainty, because we cannot verify whether the extrapolations from specific observations hold true for all (possible) cases.  -->
<!-- Induction thus only grants *plausible* conclusions--not certainty.  -->

Under the right conditions, MI yields valid, unbiased estimates (Schafer, 1997).
The modeling assumptions inherent to a model-based technique like MI cannot be tested.
Statistical models like MI are optimized to distinguish between effect and noise of population averages–not the individuals that make up your sample. 

FCS has no (deductive) proof, only by simulation.



### What is FCS?

[TODO: reframe as 'Why MI in the flavor of mice?' I.e., FCS is a flexible and agnostic way to specify conditionals to arrive at a joint. MICE also allows JOMO: a direct and inflexible way to arrive at a joint. With MICE, it's easy to generate synthetic values. mice is not just imputation software, but rather an 'information generating machine'. Describe imputation models, analysis models, non-response model and SMC.]

- In general, imputation methods are developed to solve missing data problems.
- What types of missing data problems are we considering in this dissertation? Incomplete samples where some---but not all---entries are missing per row or per column. Entirely missing rows would be 'unit non-response'; entirely missing columns would be unobserved variables. We are interested in data with incomplete cases/variables, or 'item non-response'.
- Missing observations in incomplete cases/variables can be filled in using an imputation model
- In the FCS framework [@vanbuuren2006], we need an imputation model for each incomplete variable 
- fcs is substantive model agnostic, and does not assume any specific joint distribution.
- while "Rubin (Ch. 5) distinguished three tasks for creating imputations under an explicit model: the *modeling* task the *imputation* task and the *estimation* task." [@vanbuuren2007, p. 221], FSC does not require these explicit tasks. Instead, FCS only requires an *imputation model* to describe how synthetic values may be generated, without explicit specification of the joint model.
- imputation models consist of the functional form of the model (e.g., stochastic regression) and imputation model predictors (i.e., other variables in the data)
- recurring questions: how do we choose an appropriate imputation model? and how do we know whether we have chosen an appropriate imputation model?
- "Imputation models bypass the need to specify $P(Y, X, R)$, though their use creates new responsibilities for substantiating its correctness for a given statistical analysis." [@vanbuuren2007, p. 222].
- That's where computational evaluation comes in!

::: {.callout}
# FCS =/= SMC
- the quality of the imputation model can only be determined/assessed based on the complete data model --> TODO: specify terminology!! --> comp eval also possible without analysis model vs statistical evaluation relies on known analysis model (prangender vandaag want imputation gebruikt voor meer analyses) fcs is substantive model agnostic
- what if that model is not (yet) available? then take the most complex possible model and check that, because if you would take the simplest model, then that would implcitly mean you are holding/constraining parameters (forcing b=0), systematically suppressing relations.
- another option would be like smc-fcs: we assume a complete data model and derive the imputations from there.
:::

::: {#fig-steps}

```{mermaid}
%%| fig-height: 7.5

flowchart TB
    inc[/"Incomplete data"/] --> miss[\"Inspect missingness"/] 
    subgraph build[" "]
      miss
      mod
    end
    miss --> mod[\"`Define imputation model(s)`"/]
    mod --> imp[[Run imputation algorithm]]
    imp --> imps["Imputations<br> $$\quad(1 \dots m)$$"]
    imps --> conv
    subgraph eval[" "]
      direction TB
      conv[\"`Inspect convergence`"/] 
      check[\"`Inspect imputed values`"/]
      conv <--> check
    end
    conv -- "✘ non-convergence" --> imp
    check -- "✘ misfit" --> mod
    check -- "✓ algorithmic convergence<br>✓ imputation model fit" --> analyses
    analyses[\"`Run analyses`"/] --> mira["Extract<br>estimates"]
    mira --> pool(["Pool estimates"])
    pool --> mipo["Pooled estimate"]
    inc@{ shape: doc}
    imps@{ shape: docs}
    analyses@{ shape: procs}
    mira@{ shape: tri}
```


<!-- - Giswerk/gebrek aan controle leidt to suboptimale estimates -->
<!-- merge edges: https://stackoverflow.com/questions/51367234/fusing-arrows-sideways-in-mermaid-diagrams -->

`\fignote{Statistical evaluation is mainly focused on the relation between the  incomplete data and pooled estimates. Computational evaluation is in every trapezoid: processes that require human intervention. In the first box, the question is: 'how can we model the missingness?', in the second box, the question is: 'do the imputation models fit the data?'. Missing in the graph: drawing inference with pooled estimate, and use external info after inspecting the missingness to involve all stakeholders (CRISP-DM steps).}`{=latex} 


Multiple imputation flowchart 

:::





### Building and evaluating imputation models

- Model building: how do we know which functional form to choose?
  - rely on defaults, tested in simulation studies 
  - assess the distribution of the incomplete variable (e.g., visual inspection)
  - ...? (maybe ad transformations?)
- Model building: how do we know which imputation model predictors to select?
  - association of incomplete variable with other variables in the data ["include an auxiliary variable if its correlation with the main variable is ≥0.5 (in absolute value) [25]" @nguyen2021]
  - association of missingness indicator with other variables in the data 
  - external input (e.g., branching patterns, or expert knowledge to correct for MNAR)
  - Take the analysis model (if available) as starting point for imputation workflow, for congeneality
  - ...?
- Model evaluation: how do we assess imputation model misfit?
  - fit of the model to the observed data
  - non-convergence in the algorithm 
  - mismatch between observed and imputed data distributions (e.g., visual inspection)
  - ...?

### Evaluating imputations

- FCS has no analytical proof --> rely on evidence from simulation studies
- has been shown that it works under many conditions
- but how to know in practice? we need assumptions
- like assumptions for GLM, we look at the data to evaluate the appropriateness of our chosen model
- some assumption checks are visual inspection, some are quantitative, some need external info
- similarly, we need to evaluate the choice of imputation models

- ideally: evaluate inferential validity, but only available in simulation studies
- in practice, the thing we need is not available by definition
- some assumptions are impossible to definitively determine (e.g. MNAR), some we can evaluate by looking at the data
- this is what this dissertation is about: computational evaluation

- In practice, you cannot validate whether the imputation model was appropriate (???), because what you would need to do so is exactly the thing you don't have: the missing part of the incomplete data, while you only have the observed part.
- The best thing we can do is to rely on a combination of assumptions/theoretical justifications/substantive expertise and indirect evidence of imputation model fit: check the observed versus imputed data distributions, and inspect the imputations for signs of non-convergence. This dissertation offers insight and tools to do so.
- The only setting where we do have the comparative truth (the true but missing values) is simulation studies. 
Imputers have to rely on simulation study results in order to choose the best imputation method for their incomplete variables. 
That's why simulation studies for imputation methodology should be comparable with one another.
Then, imputers can make a 'grounded' assessment which imputation methods may be applicable, and choose the most appropriate one. 
Example: NRI search for lower bound, so impute as 'positive behavior', but per definition wrong statistical inf, because we intentionally bias in one direction

- [in simulation study kun je waarheid maken, maar in de praktijk moet je aannames doen. die aannames zijn belangrijk, juist omdat ze niet/ten dele te bewijzen zijn]
- [what is a simulation study? add a brief methodological background section on simulation studies in statistics?]

### What is computational evaluation?

- Computational evaluation is not equal to statistical evaluation. 
Statistical evaluation is important, e.g. to quantify  bias and variance of estimators/procedures, but not the focus of this dissertation.
- Computational evaluation is something to check *conditional on* acceptable statistical properties, and does not replace statistical evaluation. 
- Computational evaluation is originally from information retrieval research [see e.g. @wiki]
- The practice of computational evaluation combines *effectiveness* (i.e., did the process yield the correct result, e.g., accuracy, precision, recall, etc.), *system quality* (e.g., speed, coverage of all possible results, etc.) and, importantly, *user utility* (e.g., happiness, productivity, etc.) [@manning2008]
- The term 'computational evaluation' is not specifically defined for the field of statistics, but it could be useful. 
- One definition of computational evaluation is "to determine the quality of solutions attainable" [@dudek2004] 
- Translated to statistics, a model may be statistically valid, but not fit for purpose. For example, computations that take too long to converge for use in a production pipeline/real-time use in patient care.
- Another definition of computational evaluation is "the process of ensuring an algorithmic solution is a good one: that it is fit for purpose" [@curzon2014].
- We can use this when a statistical process has different purposes and thus different usability/requirements. 
For example, calculating uncertainty at the level of a single case, a sample, or inferring to a population.
- Since the term is not defined for statistics, there is no definition to use for the computational evaluation of imputation methodology.
- In this dissertation, I want to plant a flag: how can we define 'computational evaluation' in the context of imputation methodology?
- Working definition: *systematic assessment of the algorithmic behavior and data-level outputs of imputation procedures*, focused on: algorithmic stability (convergence, failure modes), sensitivity to modeling choices, plausibility and coherence of imputed values, and usability for applied researchers. 
- These aspects are often already taken into account by imputers, but I try to systematize and operationalize these aspects that are often informal or *ad hoc*. 

Dimensions of computational evaluation: e.g. inferential validity, algorithmic behavior, data-level plausibility, user utility, ...

<!-- - which of these is the problem when we use PMM with parse data? on the one hand, it might lead to imputations that are so close that the CI is too narrow and confidence validity is ar risk. on the other hand, it might lead to implausible imputed values for cases with missingness in the sparse region. -->

- verification = "confirmation, through the provision of objective evidence, that specified requirements have been fulfilled" (ISO/IEC/IEEE) ... "process of providing objective evidence that the system, software, or hardware and its associated products conform to requirements (e.g., for correctness, completeness, consistency, and accuracy)" (IEEE) ... "Verification in a life cycle context is a set of activities that compares a product of the life cycle against the required characteristics for that product. ... The system has been built right. ... A system could be verified to meet the stated requirements, yet be unsuitable for operation by the actual users." (IEEE)
- validation = "confirmation, through the provision of objective evidence, that the requirements for a specific intended use or application have been fulfilled" (ISO/IEC/IEEE) ... "Validation in a system life cycle context is the set of activities ensuring and gaining confidence that a system is able to accomplish its intended use, goals, and objectives. The right system has been built. Validation demonstrates that the system can be used by the users for their specific tasks." (IEEE)
- verification asks whether the system has been built right whereas validation asks whether the right system has been built


### Why should we care about computational evaluation of imputation methodology?

- What do we care about? bias/coverage/variance, empirical relevance/plausibility, software-engineering nitpicks like speed/convergence/fault rate.
- Computational evaluation goes beyond statistical evaluation.
- In this dissertation I will use computational evaluation as a suite of tools to investigate the domains of algorithmic stability, data plausibility, and user burden, next to inferential validity.
- Even if we know the process yields correct inferences, we can evaluate the usability/appropriateness/plausibility of the solution. For example, one question to consider is: do the imputed values lie within the range of the observed values?
- Imputation theory does not require the imputed values to be plausible. 
- Rubin [-@rubin1976; -@rubin1987] developed the methodology to get correct population-level inferences from incomplete samples, not at the individual case-level within the sample. [TODO: consider adding that since then, data analysis has changed]
- Bluntly stated: imputation pragmatists/purists don't care for the imputed values themselves, as long as they result in statistically valid inferences.
- But what are the limitations of purely inferential metrics (bias, coverage)? e.g. imputing negative values for age, using PMM for a variable with a sparse region (yielding 'unrealistic' imputed values), or evaluating imputed values of non-converged chain (?), or not knowing the analysis model up front so congeniality is at risk (???) -> PPC would show problem with sparse region
- And vice versa: when might the plausibility of imputed values mislead a data analyst into believing the imputations are statistically valid? when 'looking reasonable' is confused with 'being consistent with the data-generating process', such as single imputation (stoch reg), or imputing under MCAR assumption when there is a MAR mechanism, or destroying the joint distribution of the data while preserving the marginals. [TODO: figure out the difference with Gerko's interpretation of joint occurrence (cross tabs) within conditional distribution; should we make an algorithm to show user which combinations of bivariate obs only occur in imp? Not now!]
- Statistical evaluation of bias and coverage is not possible in practice, only in simulation studies.
- Stat evalualtion may not possible IRL? Bias and confidence interval coverage are impossible to determine without a comparative truth (although PPC and bootstrapping allows us to compare with respect to observed values).
<!-- - statistical evaluation vs computational evaluation: does one come before the other? not necessarily. -->
- [TODO: rephrase as ability to explain differences between observed and imputed data (e.g., lower means in boys data), not as reverse engineering but as statistical evaluation of the data --> then computational eval may be needed to explain differences?] Computational evaluation means not just taking the answer as-is, but tracing back the solution of an algorithm to its origin/back-tracing the procedure: you should *always* be able to retrieve the chain of the imputation procedure. E.g., implausible or impossible imputed values can yield valid statistical inference, but are we satisfied with the way that the imputation were obtained?


<!-- Gerko: "How does computational evaluation differ from statistical eval? IMO it goes beyond it -> e.g. plausible imputations, tracing back the black box to some extend instead of just taking the answer - even if it is correct. You can give examples here wrp nonconverged solutions that yield plausible inference, implausible values that yield valid inference, etcetera."
<!-- Example: {irmi} package by mattias temple yields unbiased estimates, but coverage of 0 because variance was 0 -->
<!-- Gerko: "Would be nice to also consider that taking just a single imputed solution can not be distinguished from another sinle imputed solution, and that the monte carlo variance of the solution should be studied. Therefore it goes way beyond the statistical properties of the imputation methodology." -->
<!-- Hanne:  even if you get the same results based on different SIs, you want to be able to tell which one you should use, that leads us to the need for MI -->
<!-- Gerko: "I think it would be nice to relate it to other stuff/metrics that we value, such as bias/coverage/variance, empirical relevance/plausibility, software-engineering nitpicks like speed/convergence/fault rate." -->

::: {.callout}
# Examples of stat but not comp
- polynomials (ignoring the U in imps could still get stat val)
- example: psych study with baseline, some of which are missing; baseline, intervention, follow-up --> mice would inflate baseline and follow-up if both are incomplete; statistical evaluation will not show this (unbiased, cov ok, etc.); over-imputation or postpred check would show diff distr for obs and imp; circularity should be removed: follow-up should not predict baseline
- fcs iterations not stable, but estimand ok (in simulation study true value) --> conv at cell level??
- PMM with part of sample space with too little donors
:::

### What is the scope of this dissertation?

**RQ: what tools are required to assess the quality of data-generating algorithms? in the development, evaluation, and application.**

- is the status quo sufficient? no, statistical evaluation is required, but not sufficient.
- how Rubin developed the methodology works in theory, but is not enough for contemporary practitioners
- can we use existing tools or do we need more? and how? no, not all are present in the imputer's toolbox. I contribute to the new status quo ("verleggen" = generalizable scientific contribution of this dissertation) 
- unifying how imputations are generated and evaluated (??)
- lifecycle of imputation methods: method is theorized, method is implemented in software, method is evaluated in simulation study, method is put to scrutiny by scientific community, method is applied in practice, method is evaluated in real-world data.
- two types of evaluation: statistical properties of the method in simulation (e.g., bias and confidence-validity; statistical evaluation) & practical application (computational evaluation).

[TODO: remove what I will not address! Maybe add a sentence 'You might expect me to go into...' In the category non-response: just item, and M(C)AR mech. In the category imputation algo: just FCS. In the category substantive models: just GLM for stat inference.] In this dissertation, we will focus on missing data problems with item non-response and MCAR/MAR missingness mechanisms. Unit non-response and MNAR missingness mechanisms are outside of the scope of this dissertation. The missing data method under consideration is primarily multiple imputation by chained equations (MICE), but some results may generalize to other iterative (FCS) imputation algorithms. We do not cover joint modeling (JOMO) imputation or non-imputation missing data methods. Although MICE is not SMC-FCS, we will restrict the analysis of scientific interest mostly to GLM family of models. The applications are aimed at statistical inference (not prediction or causal inference).

### Chapters

1. Evaluation paper: evaluating (new) imputation methods via simulation studies/how we know FCS works 
2. Convergence paper: evaluating imputation algorithm quantitatively
3. Software: choosing/evaluating imputation models in practice


## Introducing chapter 1: evaluation

- **RQ: what to look out for when designing simulation studies for the evaluation of imputation methodology?**/How can computational evaluation criteria be systematically incorporated into simulation studies for imputation methodology?

- published as @oberman2024, cited 27x according to Google Scholar (of which 11x CrossRef)
- based on literature review @liu2023

- How can we trust simulation studies for imputation methods?
- How can we make them compatible & fair?

REF: Morris ADEMP: Using simulation studies to evaluate statistical methods, pre-registration template, FIMD, Introduction to statistical simulations in health research?, Pitfalls and Potentials in Simulation Studies?


## Introducing chapter 2: convergence

- **RQ: FSC is iterative, so convergence *should* be a requirement for valid inference, or is it?**/How does non-convergence in iterative imputation affect inferential validity, and how can non-convergence be diagnosed? 

- simulation study, published as pre-print (11x cited according to Google Scholar), presented at ICML

- even without convergence, we can get valid inference at sample/population level, but not at individual cell level
- imputation stays a custom job: convergence of imputation is conditional on imputation model
- today, imputation are not generated for 1 analysis prob, but all analyses that may be fitted on the data --> all imputation models should be compatible and congenieal --> more 'omvattende' imputation models required
- liefst wil je analysemodel parameters tracken, daarom is computational evaluation zo belangrijk

- Can we quantify the convergence of the MICE algorithm?

REFs: Graphical and numerical diagnostic tools to assess suitability of multiple imputations and imputation models, Diagnostic checking of multiple imputation models, Convergence Properties of a Sequential Regression Multiple Imputation Algorithm, FIMD, Diagnostics for multivariate imputations

## Introducing chapter 3: `ggmice`

[TODO: add poster and 'age old adage' picture vs 1000w.]

This chapter introduces the software `ggmice`: an R package for building and evaluating imputation models. The `ggmice` package offers tools for the visualization of incomplete data, imputation models, and imputed data. 

### Indicators for adoption: how is the software already used?

- software published on Zenodo [@oberman2022], with open code and bug reporting via GitHub
- peer-review not formal via journal, but via Issues and Pull Requests on GitHub (10x external Issue, 13x external PR)
- impact: citations (5x according to Google Scholar), CRAN downloads (852x/month; 21k total) and conda downloads (100+/month; https://anaconda.org/channels/conda-forge/packages/r-ggmice/overview)

### Design principles

- `ggmice` is compatible with the popular R packages `mice` (for imputation) and `ggplot2` (for flexible data visualizations based on the 'grammar of graphics' framework). 
- `mice` imputation models are substantive model agnostic: flexible, var-by-var definition of imputation model. `ggmice` supports this.
- Grammar of Graphics = layered plots [@wilkinson2012] --> `ggmice` produces `ggplot` objects: layered, easily adjustable

- Assumptions: imputation model should fit the observed part, imputation model should be congenial --> providing tools to see observed and imputation side-by-side/evaluation of the distribution of observed and imputed data

### Limitations

- make explicit what these plots/analyses can and cannot tell us: imputation models should fit the observed part of the data, but no way to determinately state miss mech.
- visualization =/= objective truth --> good imputation model fit/plausible values is not a guarantee for statistical validity
- visualization is not a replacement of evaluation but support, also check metrics such as FMI
- interpretation remains subjective (e.g. convergence, MAR assumption, etc.)

### What does ggmice aim to solve?

The `ggmice` package aims to solve the following problems:

 - no direct graphical comparisons between incomplete and imputed data; side-by-side presentation of the data before and after imputation; Visualizations with `ggmice` can be made to inspect marginal and joint distributions of incomplete variables side-to-side to compare incomplete and imputed data
 - no previously existing methods for visualizing `mice` imputation models; complex imputation models are easier to review through visualization (as opposed to console prints/excel exports of predictor matrix)
 - existing plotting tools for `mids` objects not easily editable/publication-ready; Visualizations with `ggmice` can be extended and adapted with tidyverse functions to create publication-ready graphs of imputed data (reflecting the uncertainty due to missingness).
 - unified toolbox for the development and evaluation of imputation models

### Existing solutions

The `ggmice` package fills a hiatus in the software landscape, with other packages generally supporting either `mice` or `ggplot2`, but not both. 
Table REF provides an overview of R packages that offer visualization tools for incomplete and imputed data. In short: with `mice` you cannot plot incomplete data distributions, all other packages lack support for `mice`-imputed data, and visualization of `mice` imputation models did not exist at all prior to `ggmice`.

```{r}
#| echo: false
# library(kableExtra)
packages <- read.csv("pkg_table.csv")
knitr::kable(packages, col.names = c("", names(packages)[-1])) 
```

<!-- https://cranlogs.r-pkg.org/badges/ggmice -->
<!-- https://naniar.njtierney.com/reference/gg_miss_upset.html -->
<!-- https://github.com/njtierney/naniar -->
<!-- https://statistikat.github.io/VIM/articles/VisualImp.html -->
<!-- https://github.com/statistikat/VIM -->
<!-- https://github.com/IQSS/amelia -->

::: {.callout}
## TODO

- write table footnote to explain all aspects of the table. i.e., these are only diffs. the table does not include things that all packages can --> maybe as appendix? 
- explain the purpose of the table: I'm both player and referee. I've got skin in the game, but I made the rubric. Someone else might evaluate/value different aspects.
- also add number of downloads: `ggmice` has 800 monthly compared to `VIM` 13k and `naniar` 22k
- make 'Imputation models' a separate section?
:::


### Software development

- Open Source Software development/FAIR --> R, GitHub, CRAN, Zenodo, ...
- CRAN downloads
- GitHub stars and issues
- StackOverflow mentions/questions
- feedback at conferences
- citations of the package (Zenodo reference)

### Future research & development

- visualizing `mira` objects
- add posterior predictive check plots & over-imputation
- quantifying uncertainty at pattern/cell level
- tools for sensitivity analyses: inducing MNAR in observed part of the data and evaluating effects on the imputations/inference
- What data-level diagnostics are informative for assessing imputation model adequacy in practice (i.e., when inferential validity cannot be evaluated)?
- pre vs post imputation facilitator functies om te checken of er nieuwe verbanden in zitten (kruistabel)


<!-- *Wat is het probleem* <br>  -->

<!-- - Visualization of incomplete data, imputation models, and imputed data.  -->
<!-- - Data exploration for building imputation models: inspect marginal and joint distributions of incomplete variables  -->
<!-- - Evaluation of 'applicablity' imputation models.  -->
<!-- - Visualization of uncertainty due to missingness after imputation.  -->

<!-- *Welke methoden bestaan al* <br> -->

<!-- - `naniar` for incomplete and imputed data (but not `mice`) -->
<!-- - `VIM` for incomplete and imputed data (but not `mice`) -->
<!-- - `mice` for imputed (but not `ggplot2`) -->
<!-- - 'artisinal' code solutions -->
<!-- - print of excel export for imputation models -->

<!-- *Wat zijn de voor- en nadelen van de huidige methoden* <br> -->

<!-- - missing visualization tool compatible with both incomplete data and `mice` imputations: `mice` lacks incomplete data, other packages lack support for `mice`-imputed data -->
<!-- - missing visualization tool for imputation models (i.e., `pred` and `meth` objects) -->

<!-- *Welke nadelen/problemen beoogt de nieuwe aanpak op te lossen* <br> -->

<!-- - no methodology for visualizing `mice` models -->
<!-- - plotting tools for `mids` objects not easily editable/publication-ready -->
<!-- - no direct comparison tool for `mice` between incomplete and imputed data -->

<!-- *In welke opzicht vult de nieuwe methode een lacune* <br> -->

<!-- - unified solution for: missingness, imputation models, imputations -->
<!-- - direct comparison between data before and after imputation with `mice` -->
<!-- - complex imputation models are easier to review through visualization (as opposed to console prints/excel exports of predictor matrix) -->

<!-- *Op welke principes is de nieuwe aanpak gebaseerd* <br> -->

<!-- - Grammar of Graphics: layered plots [@wilkinson2012] -->
<!-- - Open Source Software development/FAIR -->
<!-- - evaluation of the distribution of observed and imputed data -> imputation model should fit the observed part, imputation model should be congenial -->

<!-- *Hoe zijn deze principes in de software vertaald* <br> -->

<!-- - `ggmice` produces `ggplot` objects: layered, easily adjustable -->
<!-- - GitHub, CRAN, Zenodo, ... -->
<!-- - providing tools to see observed and imputation side-by-side -->

<!-- *Welke onderliggende aannames zijn hierbij gebruikt* <br> -->

<!-- - make explicit what these plots/analyses can and cannot tell us: imputation models should fit the observed part of the data, but no way to determinately state miss mech. -->

<!-- *Hoe kun je de huidige en nieuwe gereedschappen het beste met elkaar vergelijken* <br> -->

<!-- - data types (e.g. `mids`) -->
<!-- - how 'publication-ready' the visualizations are (e.g. number of lines of code needed* <br>) -->
<!-- - number of functions* <br>  -->
<!-- - number of downloads* <br> `ggmice` has 800 monthly compared to `VIM` 13k and `naniar` 22k -->

<!-- TODO: table with visualization types in all packages (compare `VIM`, `naniar`, `amelia`) -->


<!-- |                   Functionaliteit                  |  ggmice  |        VIM       |  naniar  |   amelia  |   mice (lattice)   | -->
<!-- |:--------------------------------------------------|:--------:|:----------------:|:--------:|:---------:|:------------------:| -->
<!-- | Visualisatie van de missingness                    | $\checkmark$       | V                | V        | (beperkt) | V                  | -->
<!-- | Visalisatie van het imputatie EN nonresponse model | V        | X                | X        | X         | X                  | -->
<!-- | Imputatie diagnostiek (visueel) + evt in cijfer?   | V        | V (gedeeltelijk) | X        | V         | V                  | -->
<!-- | Ondersteuning mids en andere data typen            | V        | X                | X        | X         | V (alleen lattice) | -->
<!-- | Grammar of Graphics                                | V        | X                | V        | X         | X                  | -->
<!-- | Aanpasbaarheid                                     | hoog     | laag             | hoog     | laag      | laag               | -->
<!-- | Tidyverse compatible?                              | volledig | beperkt          | volledig | beperkt   | geen               | -->
<!-- | Meteen klaar voor publicatie?                      | hoog     | middel           | hoog     | laag      | laag               | -->


<!-- *Levert het nieuwe gereedschap in de praktijk betere resultaten* <br> -->

<!-- - visualization =/= objective truth -> good imputation model fit/plausible values is not a guarantee for statsistical validity -->
<!-- - visualization is not a replacement of evaluation but support, also check metrics such as FMI -->
<!-- - interpretation remains subjective (e.g. convergence, MAR assumption, etc.) -->

<!-- *Wat vinden toekomstige gebruikers van het gereedschap* <br> -->

<!-- - StackOverflow, GitHub issues, feedback at conferences -->

<!-- *Wat zijn relevante indicatoren voor adoptie* <br> -->

<!-- - CRAN downloads -->
<!-- - GitHub stars and issues -->
<!-- - StackOverflow mentions/questions -->
<!-- - feedback at conferences -->
<!-- - citations of the package (Zenodo reference) -->

<!-- *Welke richtlijnen en advies gelden bij gebruik* <br> -->

<!-- - not a replacement of evaluation but support -->

<!-- *Wat zijn beperkingen van de methode* <br> -->

<!-- - interpretation remains subjective (e.g. convergence, MAR assumption, etc.) -->

<!-- *Wat zijn vragen voor nader onderzoek* <br> -->

<!-- - visualizing `mira` objects -->
<!-- - add posterior predictive check plots & over-imputation -->
<!-- - quantifying uncertainty at pattern/cell level -->
<!-- - tools for sensitivity analyses -->
<!-- - What data-level diagnostics are informative for assessing imputation model adequacy in practice (i.e., when inferential validity cannot be evaluated)? -->
<!-- - Do we get better imputations in practice? -->

\newpage

# Chapter 1: evaluation

[TODO: link/insert]

\newpage

# Chapter 2: convergence

[TODO: link/insert]

\newpage

# Chapter 3: ggmice

[TODO: link/insert]

\newpage

# Discussion

<!-- Goal: -->
<!-- consolidate contributions -->
<!-- interpret, not restate—findings -->
<!-- acknowledge limitations -->
<!-- articulate future research agenda -->

## Conclusion

- Computational evaluation of imputation methodology is never finished.

- The main findings in this thesis are...
  - Defining and testing new imputation methods through simulation studies requires careful consideration of the simulation design and evaluation to make imputation methods comparable across studies. 
  We need simulation studies to know imputation methods work. 
  But we cannot be certain they actually *did* work whenever we apply them. The missing part of the data is by definition not available, so we cannot definitively state anything about the imputation model fit to the unobserved part of the data. Moreover, there's untestable assumptions (e.g., MNAR) and lack of clear diagnostic cut offs (e.g., convergence).
  - Non-convergence is hard to diagnose, and typical thresholds to evaluate non-convergence may not be applicable to FCS algorithms: MICE reaches stable output before non-convergence metrics would indicate so. 
  There is a mismatch between theoretical convergence criteria and practical algorithmic behavior. 
  That's why we still have to rely on visual inspection.
  - Visual inspection aids imputers in developing and evaluating imputation models, because formal tests are not available. 
  The R package `ggmice` offers tools for the computational evaluation of the imputations models that were previously unavailable.

- Recommendations for fellow missing data methodologists:
  - design simulation studies structured and reproducible (e.g., report design choices, publish code)
  - talk to your intended audience (i.e., applied researchers), or at least check what they are struggling with (e.g., StackOverflow)
  - ...?

- Recommendations for fellow imputers:
  - think before you code 
  - look at the data, plot the data
  - ...?

## Limitations

- Computational evaluation is not a substitute for thinking. It may give a false sense of certainty, e.g. against MNAR mechanism (untestable assumption).
- Instead, use computational evaluation as complement to (not a replacement for) substantive knowledge and sensitivity analyses. Use expert insight


- This entire dissertation relies on the R language and centers the `mice` package [TODO rephrase as not a flaw: I focus on a dominant ecosystem, and concepts generalize beyond].
  - While this set-up is very popular, the data science landscape has expanded and is ever evolving
  - Many machine learning and deep learning methods are not (yet) implemented as imputation models in `mice`. These might be able to better approximate the posterior predictive distribution of the missing values than `mice` methods, but research should show that still.

- ~~I haven't tested whether the tools I developed actually improve the imputations of `mice` users~~ This dissertation evaluates tools and diagnostics at the methodological level; assessing their causal impact on user behavior and imputation quality is not feasible to systematically test.
  - GitHub issues and StackOverflow are indication
  - Onmogelijk om evidence-based conclusies te trekken, maar misschien wel evidence-informed?

- The visualizations implemented in `ggmice` are generally better suited for numeric data, as opposed to categorical variables (or even open text field data, which is not supported by `mice` currently). Future research on: associations of cat vars with miss indicators, and convergence parameter other than SD of imputed values.

::: {.callout}
- TODO: rephrase `haven't's into future research (positive phrasing: reframe as scope conditions, not shortcomings)
- maybe future: computational evaluation of variable importance within imputation models
:::

## Outlook

- This chapter ends with a future perspective...

- Missing data remains a problem indefinitely. I expect imputation methodology to be developed and applied for the foreseeable future. But the way we deal with data might change.
- With the rise of machine learning and deep learning methods (or "AI"), evaluation will become ever more important. Novel methods are being developed under a prediction framework, not statistical inference framework: ignoring the uncertainty in the estimates, and thus not incorporating between-imputation variance. This leads to too narrow CIs, and too low p-values. Which, in turn, causes spurious results.
- AI models should propagate uncertainty!

- trust in science requires honest reflection of uncertainty in our analyses
- be open about uncertainty, and limitations of scientific studies 
- publish null results (e.g., registered reports or pre-registrations)
- share data and code with other scientists, and the public
- publish open access, educational materials too!

- change the way we evaluate science: this dissertation is an example of the new recognition and rewards vision. it includes software as an actual chapter, not something extra.
- I hope to inspire others to reflect on their own view of what scientific output is. And most of all, I hope to set an example for other PhD candidates to do the same

- imputation is not always the way to go: missingness can tell you about mismatch between data collection method and lived experience --> positionality??

\newpage

# References

::: {#refs}
:::

\newpage

# CV

[TODO: write]
