# Introduction {#sec-intro .unnumbered}

```{r}
#| echo: false
source("_common.R")
```

<!-- ### TL;DR (or: the quality of the solution) -->

```{r}
#| message: false
#| warning: false
#| include: false
library(ggplot2)
library(ggmice)
load("data/dat_xy.RData")
```

<!-- Algorithms are everywhere.  -->

This dissertation is about algorithms and the quality of their output. 
I aim to investigate how one specific type of data-generating algorithm can be evaluated for use in statistical inferences. 
More specifically, I focus on imputation algorithms: algorithms with the purpose of generating plausible data values for information that is not observed, based on a model of some observed information. 
This type of algorithm is popular for solving missing data problems in practice.  

<!-- The data-generating algorithms in question are 'imputation' algorithms. -->
<!-- - The algorithm central to this dissertation is MICE: multivariate imputation by chained equations. The MICE algorithm is developed to generate 'imputed values': synthetic substitutes to replace missing values in incomplete data, based on observed information.  -->

<!-- - MICE is an iterative algorithm. It produces imputations on a variable-by-variable basis. After a full pass through the data, the algorithm cycles back to the first column to replace the imputed value from the first iteration with an updated entry. -->

<!-- generate several completed versions of an incomplete dataset, which allows data analysts to handle the completed datasets as if they were completely observed.  -->

Missing data problems are timeless, but technological advances have made treating missingness easier than ever [REFs?]. 
Nowadays, a single line of code can seem to solve your missing data problem. 
But is that solution really what you would expect? 
And does that solution give you the correct statistical answer?
That is what my dissertation is about.
I want to contribute methodology that facilitates drawing valid inferences from incomplete data. 

Missing data are ubiquitous and cannot be ignored without risking bias.
Missing data occur often in scientific disciplines with human subjects, such as the social sciences and biomedical research [REFs?].
A participant in a study might, for example, drop out of a longitudinal trial or refuse to answer a sensitive question, resulting in gaps in the research data.
These gaps complicate data analysis and may lead to biased and invalid study results.
That is where imputation comes into play.

Imputation algorithms are designed to 'impute' (i.e., fill in) missing values in incomplete datasets. 
Imputation is a popular and flexible procedure that separates the missing data problem from the analysis of scientific interest. 
Imputation allows data analysts to first solve for the missingness by imputing the missing values, to then be able to analyze the completed data as if the data were completely observed. 
If done several times in parallel, imputation can yield estimates that are both unbiased as well as confidence valid in a broad range of scenarios.
This practice of drawing inference by replacing missing values with multiple synthetic substitutes is called 'multiple imputation' (MI). 
@fig-imp provides an example of multiple imputation, where a missing value is replaced several times with plausible substitutes. 
<!-- TODO: explain figure. [Gerko: Ik zou hier wel echt kapitaliseren op deze afbeelding. Je kunt heel mooi uitleggen dat de waardes voor X hier ontbreken en dat Y wel is geobserveerd en dat we de relatie tussen Y en X juist gebruiken om deze waarden te schatten. Je kunt ook heel mooi uitleggen dat iedere waarde voor X onzeker is en dat je daarom juist de Monte Carlo variatie hier tussen mee zou willen nemen] -->
The variability between imputations will reflect the uncertainty due to non-response in the final estimate.

```{r}
#| eval: false
#| include: false
knitr::kable(dplyr::arrange(dat_xy, Y))
```

::: {#fig-imp}

```{r}
#| layout-ncol: 2
#| fig-height: 3
# dat_xy[6, 2] <- 5.5
# TODO: add caption
ggmice(dat_xy, aes(X, Y)) + 
  geom_point() + 
  theme_classic() +
  theme(legend.position = "none") +
  annotate(
    geom = "curve", x = 2.5, y = 7.5, xend = 1.8, yend = 6, 
    curvature = .1, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate("text", x = 3.6, y = 8, label = "missing value")
a <- mice::mice(dat_xy, m = 3, method = "norm", seed = 1, print = FALSE)
ggmice(a, aes(X, Y)) + 
  geom_point() + 
  scale_x_continuous(limits = c(1.7, 9), breaks = c(2, 4, 6, 8)) +
  theme_classic() +
  theme(legend.position = "none") +
  annotate(
    geom = "curve", x = 5, y = 7.5, xend = 4.5, yend = 6, 
    curvature = .3, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate(
    geom = "curve", x = 6, y = 7.4, xend = 5.8, yend = 6, 
    curvature = .2, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate(
    geom = "curve", x = 7.8, y = 7.5, xend = 8.3, yend = 6, 
    curvature = -.3, arrow = arrow(length = unit(2, "mm"))
  ) +
  annotate("text", x = 6.5, y = 8, label = "imputed values")  
```

<!-- `\fignote{This is plain text turned as LaTeX caption.}`{=latex} -->

Multiple imputation of a missing value

:::

<!-- For data analysts, missing data are a nuisance. Missingness stands in the way of unbiased and confidence valid statistical inferences. 
Whenever there is just one missing value present in a dataset, extrapolation from the sample to the population may be hindered. 
Firstly, because incomplete data always require some treatment for the missingness. 
Even methods that seemingly ignore the missing data, such as complete case analysis, are a type of missing data treatment. Missing data treatment can severely impact the accuracy and validity of any subsequent analyses. 
Therefore, missing data methods should be applied carefully. 
Secondly, because many missing data methods lead to biased estimates if the incomplete data are in any way different from the complete data one would have liked to obtain. 
For example, when drop-out in a clinical trial is related to the severity of side-effects from the treatment.  -->

<!-- Whenever missing data cannot be prevented, the only way forward is to treat the missing data problem. Missing values force data analysts to treat the missingness in some way.  -->

<!-- Missing data may be widespread, but so are missing data treatment methods.  -->

<!-- Imputation is a convenient and popular missing data method.  -->

Although imputation is widely supported by software and many implementations of the method have convenient defaults, generating good imputations is not a trivial task.
**How can we know when to trust the output of imputation algorithms?**
There are three approaches: analytical proof, experiments, and applications.
The ideal route is to study the formal statistical properties of the method via analytical derivations.
Unfortunately, this is often impossible for imputation algorithms, many of which do not have formal proofs.
Instead, the de facto standard is to estimate their empirical performance via simulation studies.
These studies, however, are not standardized and are therefore hard to compare.
Moreover, simulation study results might not always translate to applied practice.
For real-world applications, there is no systematic framework for evaluating the operational utility of imputation algorithms, such as their algorithmic convergence.
Ergo, evaluating whether imputations should be trusted remains challenging.
<!-- TODO: add 'subjective to some extend'? -->

  <!-- - Theoretical properties: analytically (often impossible) -->
  <!-- - Empirical performance: via simulation studies ('in silico'; the standard, but not standardized) -->
  <!-- - In practice/applied use: evaluate output, sensitivity analyses, ...? (very ad hoc, not systematic) -->
  
In this dissertation, I set out to investigate how the methodology for developing, applying, and evaluating data-generating algorithms may be standardized. 
I thereby focus on the imputation procedure 'MICE' [@vanbuuren1999] as implemented in the R package `mice` [@vanbuuren2011].

I *study* the evaluation of imputation methodology because current evaluation practices (such as computational evaluation via simulation studies, and graphical evaluation via diagnostic plots) are fragmented and may be insufficient. 
I *contribute to* the evaluation of imputation methodology by bridging inferential evaluation and data- or algorithm-level diagnostics for imputation workflows.

<!-- - Missing data may be widespread, but so are missing data treatment methods. -->
<!-- - Imputation = filling in missing values to fix the missing data problem. -->
<!-- - The nuisance of missing data is often relieved using imputation methods.  -->
<!-- - To 'impute' means to fill in, so 'imputation' is the practice of replacing missing values with synthetic substitutes.  -->
<!-- - imputation splits the missing data problem and the analysis of scientific interest: first fill in the missingness to be able to analyze the completed data as if they were completely observed -->
<!-- - Imputation is easy with software defaults [e.g. in `mice`, @mice] -->
<!-- - But good imputation is very hard still. -->
<!-- - How do we know if we have a good imputation method? -->
<!-- - In this dissertation, I study computational evaluation of imputation methodology. -->

<!-- Missing data problems are timeless. Technological advances have made imputation fairly easy to apply, but how to apply it *well*? -->

<!-- [TODO: add take-home message in 1-2 sentences without jargon?] -->

<!-- $$Y = \{Y_{obs}, \dot{Y}_{\text{mis}}\}$$ -->
