# Research {#sec-research .unnumbered}

```{r}
#| echo: false
source("_common.R")
```

### What to expect in this dissertation? (or: structure & scope)
<!-- Application/chapters -->

In this dissertation, I investigate how imputation workflows with `mice` [@vanbuuren2011] can be evaluated.
I frame imputation processes as sociotechnical systems, based on a system engineering conceptual framework.
This allows us to apply the verification and validation (V&V) standards, as defined for software systems [@ieee2025].
I operationalize V&V through computational and graphical evaluation tools. 
Verification of imputation methodology mostly concerns computational evaluation tools, such as simulation.
Validation requires both computational as well as graphical evaluation tools, to assess the operational utility of imputation workflows.
Each of the three substantive research chapters in this dissertation covers one of these aspect: computation evaluation for verification, computational evaluation for validation, and graphical evaluation for validation, respectively.
<!-- the use of computational and graphical evaluation tools for imputation .  -->
<!-- This disseration consists of three substantive research chapters. -->

<!-- Chapter 2 is dedicated to  -->
<!-- Chapters: -->

<!-- 1. Evaluation paper: evaluating (new) imputation methods via simulation studies/how we know FCS works  -->
<!-- 2. Convergence paper: evaluating imputation algorithm quantitatively -->
<!-- 3. Software: choosing/evaluating imputation models in practice -->

<!-- Imputation with `mice` is a sociotechnical system. -->

<!-- And there should be interoperability with other systems an imputer might use in their imputation workflow, such as `ggplot2` [@wickham2016] for data visualization, and statistical analysis packages such as `stats` [@rcoreteam2025]. -->
<!-- That is where V&V comes in. -->

<!-- - Verification of statistical properties via formal analysis (e.g. proofs) and/or simulation -->
<!-- - Validation of operational utility via graphical and computational evaluation -->

<!-- #### Verification --> 
<!-- Simulation studies are verification. -->
<!-- Assessing whether the system yields correct (known) output when a specific (known) input is supplied. -->
<!-- We want to check whether the system (or the method in this case) conforms to the requirements that we set (e.g. accuracy).  -->
<!-- Statistical evaluation is the default to evaluate the inferential validity.  -->
<!-- Examples are unit tests for software and performance assessment of methods via simulation.  -->
<!-- For imputation methodology, verification is what we do with simulation studies and with unit tests in software.  -->
<!-- Verification via simulation is done, but not standardized.  -->
<!-- Verification via testing is mostly done as unit tests, but not as integration tests. -->


<!-- #### Validation -->
<!-- In simulation studies we can construct the "truth", but in practice we have to rely on assumptions. Those assumptions are crucial precisely because they cannot be (fully) proven. -->
<!-- In practice, you cannot validate whether the imputation model was appropriate, because what you would need to do so is exactly the thing you don't have: the missing part of the incomplete data, while you only have the observed part. -->
<!-- The thing we need is not available by definition. -->
<!-- In practice, the best thing we can do is to rely on a combination of assumptions/theoretical justifications/substantive expertise and indirect evidence of imputation model fit: check the observed versus imputed data distributions, and inspect the imputations for signs of non-convergence.  -->
<!-- Some assumptions are impossible to definitively determine (e.g. MNAR), some we can evaluate by looking at the data. -->
<!-- Inferential validity cannot be established. -->
<!-- This dissertation offers insight and tools to do so. -->
<!-- Validation tools assess operational utility. -->
<!-- This disseration is mostly concerned with valitation methodology via computational and graphical evaluation. -->

### Computational evaluation for verification
<!-- Introducing chapter 2: evaluation -->

In Chapter 2, I investigate how we can standardize the verification of imputation methodology.
<!-- what computational evaluation for verifying imputation methodology can and should look like. -->
I posit simulation studies as a computational evaluation tool to judge whether imputation strategies work experimentally.
A question underlying this chapter is: how can computational evaluation criteria be systematically incorporated into simulation studies for imputation methodology?
Other questions that come up are: how can we design simulation studies that are fair and comparable with one another, which simulation conditions to consider, which performance metrics to report, and how to document this in such a way that others can reproduce and reuse the evaluations?
I address these questions based, in part, on a literature review I was involved in as a research assistant [@liu2023]. 
This chapter focuses on methodologists (who design imputation methods and study their properties) rather then imputers (applied researchers who apply imputation models to specific datasets).
Methodologists need comparable, reproducible evaluations of imputation methods across scenarios.
Without a shared evaluation framework, claims about 'valid, unbiased estimates' might not translate to practice.
I argue that imputation methodology should be subjected to a standardized evaluation framework for simulation studies.

<!-- how computational evaluations can be designed, standardized, and interpreted to provide credible verification of imputation methodology. -->
<!-- In other words, can  -->
<!-- how we should judge, in a systematic way, whether imputation strategies work. -->
<!-- I argue that imputation methods should be subjected to a standardized evaluation framework for simulation studies. -->
<!-- How to design simulation studies that are comparable?  -->
<!-- Which simulation conditions to consider, which performance metrics to report?  -->
<!-- And how to document this in such a way that others can reproduce and reuse the evaluation? -->

<!-- The setting where we do have the comparative truth (the true but missing values) is simulation studies. -->
<!-- Imputers have to rely on simulation study results in order to choose the best imputation method for their incomplete variables.  -->
<!-- That's why simulation studies for imputation methodology should be comparable with one another. -->
<!-- Then, imputers can make a 'grounded' assessment which imputation methods may be applicable, and choose the most appropriate one.  -->
<!-- Example: NRI search for lower bound, so impute as 'positive behavior', but per definition wrong statistical inf, because we intentionally bias in one direction -->

<!-- This chapter asks how we should judge, in a systematic way, whether imputation strategies work. -->
<!-- **RQ: what to look out for when designing simulation studies for the evaluation of imputation methodology?**/How can computational evaluation criteria be systematically incorporated into simulation studies for imputation methodology? -->

<!-- - How can we trust simulation studies for imputation methods? -->
<!-- - How can we make them comparable & fair? -->

<!-- Focus on methodologists (who design imputation methods and study their properties) as opposed to imputers (applied researchers who apply imputation models to specific datasets). -->
<!-- Methodologists need comparable, reproducible evaluations of imputation methods across scenarios. -->
<!-- Without a shared evaluation framework, claims about 'valid, unbiased estimates' for MI might not translate to practice. -->
<!-- How can we standardize the verification of imputation methodology? -->

<!-- The chapter offers advice for designing simulation studies (i.e., computational evaluation) -->
<!-- - published as @oberman2024, cited 27x according to Google Scholar (of which 11x CrossRef) -->
<!-- based on literature review @liu2023. -->


  <!-- TODO: add?  -->
  <!-- Defining and testing new imputation methods through simulation studies requires careful consideration of the simulation design and evaluation to make imputation methods comparable across studies.  -->
  <!-- We need simulation studies to know imputation methods work.  -->
  <!-- But we cannot be certain they actually *did* work whenever we apply them.  -->
  <!-- The missing part of the data is by definition not available, so we cannot definitively state anything about the imputation model fit to the unobserved part of the data.  -->
  <!-- Moreover, there's untestable assumptions (e.g., MNAR) and lack of clear diagnostic cut offs (e.g., convergence). -->
  
<!-- TODO: add REFs? Morris ADEMP: Using simulation studies to evaluate statistical methods, pre-registration template, FIMD, Introduction to statistical simulations in health research?, Pitfalls and Potentials in Simulation Studies? -->

<!-- formal and empirical properties -->

### Computational evaluation for validation

<!-- ### Introducing chapter 3: convergence   -->

<!-- [TODO: add] -->
<!-- FCS imputation,  is an iterative procedure.  -->
In Chapter 3, I study how one might diagnose algorithmic non-convergence in `mice` [@vanbuuren2011].
This may be considered a type of computational evaluation for validation, since non-convergence diagnostics can flag problems in the imputation process to `mice` users.
Non-convergence may happen because MICE (as implemented in `mice`) is an iterative algorithm.

Incomplete variables are imputed one-by-one, conditional on other variables.
If these other variables are incomplete too, iteration is required.
After the first cycle through the columns, we can get better imputations based on the now-updated imputation model predictors.
In `mice`, this cycle is repeated several times until convergence is reached (by default five iterations).
Because of its iterative nature, convergence *should* be a requirement for valid inference, or is it?
This chapter investigates two questions: 1) how does convergence in iterative imputation affect inferential validity, and 2) how can non-convergence be diagnosed?
These questions are answered via a simulation study. 
<!-- , published as pre-print (11x cited according to Google Scholar), presented at ICML -->

<!-- - TODO: add that even without convergence, we can get valid inference at sample/population level, but not at individual cell level -->
<!-- - TODO: add that imputation stays a custom job: convergence of imputation is conditional on imputation model -->
<!-- - TODO: add that today, imputation are not generated for 1 analysis prob, but all analyses that may be fitted on the data -> all imputation models should be compatible and congenieal -> more 'omvattende' imputation models required -->
<!-- - liefst wil je analysemodel parameters tracken, daarom is computational evaluation zo belangrijk -->

<!-- - Can we quantify the convergence of the MICE algorithm? -->

<!-- TODO: add REFs? Graphical and numerical diagnostic tools to assess suitability of multiple imputations and imputation models, Diagnostic checking of multiple imputation models, Convergence Properties of a Sequential Regression Multiple Imputation Algorithm, FIMD, Diagnostics for multivariate imputations -->

### Graphical evaluation for validation

<!-- ### Introducing chapter 4: `ggmice` -->

<!-- TODO: add minimal example? -->
The fourth chapter introduces the software `ggmice` [@oberman2025], an R package for building and evaluating imputation models. 
The `ggmice` package offers tools for the visualization of incomplete data, imputation models, and imputed data. 
`ggmice` supports `mice` workflows by making imputation modeling choices visible and, therefore, open to scrutiny.
Compatibility with `ggplot2`'s 'grammar of graphics' [@wickham2016] makes `ggmice` graphs extendable with the familiar tools of the `tidyverse` [@wickham2019], and adaptable for publication-ready visualizations.

<!-- #### Indicators for adoption: how is the software already used? -->

Although `ggmice` has not undergone the traditional journal-based peer review, it has been vetted in practice.
The package has been released on CRAN in 2022 [@oberman2025]. 
CRAN downloads are reaching approximately 1.000x per month at the time of writing [@balamuta2026].
The accompanying Zenodo publication [@oberman2022] has been cited in seven scientific works.
Open source code and issue tracking are available in a repository on the programming platform GitHub ([github.com/amices/ggmice](https://github.com/amices/ggmice)).
There have been multiple external 'Issues' and 'Pull Requests' on GitHub, indicating that other researchers and developers have actively reviewed and improved the codebase.
<!-- - software published on Zenodo [@oberman2022], with open code and bug reporting via GitHub -->
<!-- - peer-review not formal via journal, but via Issues and Pull Requests on GitHub (10x external Issue, 13x external PR) -->
<!-- - impact: citations (5x according to Google Scholar), CRAN downloads (852x/month; 21k total) and conda downloads (100+/month; https://anaconda.org/channels/conda-forge/packages/r-ggmice/overview) -->
In short, there is active use of and involvement in `ggmice`, as indicated by regular downloads from CRAN, citations to the Zenodo record in the scientific literature, and community feedback via platforms such as GitHub and StackOverflow.

<!-- #### Design principles -->

`ggmice` is developed according to 'open source' and FAIR principles. 
The package is implemented in the open source language R, distributed via CRAN and GitHub, and archived on Zenodo.
Its evolution is driven by user feedback (e.g., GitHub issues, StackOverflow questions, conference discussions, teaching) and collaboration (e.g., code contributions, statistical consulting with external partners).

Other design principles are grounded in `ggmice`'s compatibility with the popular R packages `ggplot2` and `mice`. 
`ggplot2` offers flexible data visualizations based on the 'grammar of graphics' framework.
`mice` offers flexible imputation workflows for inference with incomplete data.
Together, these packages inform the design and functionalities of `ggmice`.

`ggmice` produces `ggplot` objects, leveraging the 'grammar of graphics' to create layered, easily adjustable plots. 
<!-- Grammar of Graphics = layered plots [@wilkinson2012] `ggmice` produces `ggplot` objects: layered, easily adjustable -->
The figure that appears is the output of a series of design decisions.
The grammar of graphics treats plots as structured ways of reasoning from data [@wilkinson2012].
Layering decomposes graphics into a small set of elements.
For example, variables are translated into visual properties (axes, color, etc.) via 'aesthetic mapping', and geometric objects determine the shape of the graph itself.
A graphic then is a combination of several elements, and layering allows for building complex figures.
<!-- Conceptually, this is a 'model' of the visualization.  -->
The user can add or change layers without breaking the entire figure, making the graphics transparent and manipulable.

`mice` imputation models are substantive model agnostic: variable-by-variable definition of imputation models. `ggmice` supports this. 
Because `mice` is substantive-model agnostic, `ggmice` offers tools to inspect whether the imputation model fits the observed data, instead of imposing particular analysis models.
Assumptions: imputation model should fit the observed part, imputation model should be congenial 
`ggmice` provides tools to see observed and imputation side-by-side, adhering to advice to evaluate the distribution of observed and imputed data [@nguyen2017].

<!-- #### What does ggmice aim to solve? -->

`ggmice` offers a unified toolbox for the development and evaluation of imputation models.
The `ggmice` package aims to solve the following problems:

- No direct graphical comparisons between incomplete and imputed data.
 <!-- side-by-side presentation of the data before and after imputation.  -->
 Visualizations with `ggmice` can be made to inspect marginal and joint distributions of incomplete variables side-to-side to compare incomplete and imputed data. 
 Without `ggmice`, it is cumbersome to see the data before and after imputation in comparable plots. 
 `ggmice` offers functions that visualize incomplete data distributions and their completed counterparts.

- No previously existing methods for visualizing `mice` imputation models.
 The structure of `mice` imputation models (i.e., predictor matrix and methods vector) are often printed to the R console as text.
 <!-- Sometimes, they may be exported to spreadsheets.  -->
 Complex imputation models are easier to review through visualization.
 `ggmice` provides imputation model plots that make complex imputation specifications easier to review and discuss.
 
- Limited chart types and design options for imputed data.
  Existing plotting tools for `mids` objects not easily editable or modifiable with different chart types.
  Visualizations with `ggmice` can be extended and adapted with `tidyverse` functions to create publication-ready graphs of imputed data (reflecting the uncertainty due to missingness).

<!-- #### Existing solutions -->

The `ggmice` package fills a hiatus in the software landscape, with other packages generally supporting either `mice` or `ggplot2`, but not both. 
@tbl-software provides an overview of R packages that offer visualization tools for incomplete and imputed data. 
In short: with `mice` you cannot plot incomplete data distributions, all other packages lack support for `mice`-imputed data, and visualization of `mice` imputation models did not exist at all prior to `ggmice`.

In particular, visualization of `mice` imputation models did not exist prior to `ggmice`.
And side-by-side visualizations of incomplete and imputed data was not supported in a unified graphical framework (nl., `naniar`'s [@tierney2023] `ggplot` graphs with missing data, versus `mice`'s `lattice` [@sarkar2008] plots with imputed data).

<!-- https://cranlogs.r-pkg.org/badges/ggmice -->
<!-- https://naniar.njtierney.com/reference/gg_miss_upset.html -->
<!-- https://github.com/njtierney/naniar -->
<!-- https://statistikat.github.io/VIM/articles/VisualImp.html -->
<!-- https://github.com/statistikat/VIM -->
<!-- https://github.com/IQSS/amelia -->

::: {tbl-software}

```{r}
#| echo: false
#| label: tbl-software
#| tbl-cap: "Differential functionality of `ggmice` versus other R packages"
# library(kableExtra)
packages <- read.csv("tables/pkg_table.csv")
knitr::kable(packages, col.names = c("", names(packages)[-1])) 
```

<!-- - write table footnote to explain all aspects of the table. i.e., these are only diffs. the table does not include things that all packages can; maybe as appendix?  -->
<!-- - explain the purpose of the table: I'm both player and referee. I've got skin in the game, but I made the rubric. Someone else might evaluate/value different aspects. -->
<!-- - also add number of downloads: `ggmice` has 800 monthly compared to `VIM` 13k and `naniar` 22k -->
<!-- <!-- - make 'Imputation models' a separate section? --> 
<!-- https://r-observatory.thecoatlessprofessor.com/downloads/compare/VIM...Amelia...ggmice -->

<!-- only includes functionality that `ggmice` offers Only differences in functionality in R packages for visualizing incomplete and imputed data  of explain the purpose of the table: I'm both player and referee. I've got skin in the game, but I made the rubric. Someone else might evaluate/value different aspects. also, these are only diffs. the table does not include things that all packages can. -->

`\fignote{This is an incomplete overview of the R package's functionalities. The purpose of this table is to display \textit{differences} in functionality based on the features offered by \texttt{ggmice}. I intend to show \texttt{ggmice}'s added value compared to existing methodology, not a universal comparison. Symbols denote high/well supported ($+$), medium/moderate support ($\pm$), and low/not supported ($-$). }`{=latex} 

:::

<!-- #### Limitations -->

`ggmice` does not pretend that visualization yields objective truth.
A good graphical fit between observed and imputed data (e.g. overlapping scatterplots) is not a guarantee of statistical validity or correct missingness assumptions.
Visual diagnostics can only suggest plausibility or highlight obvious misfits (imputation models should fit the observed part of the data).
They cannot determinately identify the missing-data mechanism.
The interpretation remains subjective (e.g. convergence, MAR assumption, etc.).
Visualization is not a replacement for, but support to statistical and computational evaluation.
<!-- Also check metrics such as FMI. -->
In short, `ggmice` is an attempt to make imputation model assumptions and consequences as transparent as possible.

<!-- - make explicit what these plots/analyses can and cannot tell us: imputation models should fit the observed part of the data, but no way to determinately state miss mech. -->
<!-- - visualization =/= objective truth -> good imputation model fit/plausible values is not a guarantee for statistical validity -->
<!-- - visualization is not a replacement of evaluation but support, also check metrics such as FMI -->
<!-- - interpretation remains subjective (e.g. convergence, MAR assumption, etc.) -->

<!-- ### Software development -->

<!-- - Open Source Software development/FAIR -> R, GitHub, CRAN, Zenodo, ... -->
<!-- - CRAN downloads -->
<!-- - GitHub stars and issues -->
<!-- - StackOverflow mentions/questions -->
<!-- - feedback at conferences -->
<!-- - citations of the package (Zenodo reference) -->

### Scope

In this dissertation, I will focus on missing data problems with item non-response and ignorable missingness mechanisms. 
You might expect me to go into other types of missingness (e.g., unit non-response, and MNAR mechanisms), but these are outside the scope of this dissertation.
I will also limit the applications in this dissertation to the general linear models framework for statistical inference.
This excludes other types of substantive models and analysis frameworks (e.g., deep learning methods, and causal inference).
The imputation method of interest in this disseration is FCS in the flavor of MICE.
However, some results may generalize to other iterative (FCS) imputation algorithms.

<!-- *Wat is het probleem* <br>  -->

<!-- - Visualization of incomplete data, imputation models, and imputed data.  -->
<!-- - Data exploration for building imputation models: inspect marginal and joint distributions of incomplete variables  -->
<!-- - Evaluation of 'applicablity' imputation models.  -->
<!-- - Visualization of uncertainty due to missingness after imputation.  -->

<!-- *Welke methoden bestaan al* <br> -->

<!-- - `naniar` for incomplete and imputed data (but not `mice`) -->
<!-- - `VIM` for incomplete and imputed data (but not `mice`) -->
<!-- - `mice` for imputed (but not `ggplot2`) -->
<!-- - 'artisinal' code solutions -->
<!-- - print of excel export for imputation models -->

<!-- *Wat zijn de voor- en nadelen van de huidige methoden* <br> -->

<!-- - missing visualization tool compatible with both incomplete data and `mice` imputations: `mice` lacks incomplete data, other packages lack support for `mice`-imputed data -->
<!-- - missing visualization tool for imputation models (i.e., `pred` and `meth` objects) -->

<!-- *Welke nadelen/problemen beoogt de nieuwe aanpak op te lossen* <br> -->

<!-- - no methodology for visualizing `mice` models -->
<!-- - plotting tools for `mids` objects not easily editable/publication-ready -->
<!-- - no direct comparison tool for `mice` between incomplete and imputed data -->

<!-- *In welke opzicht vult de nieuwe methode een lacune* <br> -->

<!-- - unified solution for: missingness, imputation models, imputations -->
<!-- - direct comparison between data before and after imputation with `mice` -->
<!-- - complex imputation models are easier to review through visualization (as opposed to console prints/excel exports of predictor matrix) -->

<!-- *Op welke principes is de nieuwe aanpak gebaseerd* <br> -->

<!-- - Grammar of Graphics: layered plots [@wilkinson2012] -->
<!-- - Open Source Software development/FAIR -->
<!-- - evaluation of the distribution of observed and imputed data -> imputation model should fit the observed part, imputation model should be congenial -->

<!-- *Hoe zijn deze principes in de software vertaald* <br> -->

<!-- - `ggmice` produces `ggplot` objects: layered, easily adjustable -->
<!-- - GitHub, CRAN, Zenodo, ... -->
<!-- - providing tools to see observed and imputation side-by-side -->

<!-- *Welke onderliggende aannames zijn hierbij gebruikt* <br> -->

<!-- - make explicit what these plots/analyses can and cannot tell us: imputation models should fit the observed part of the data, but no way to determinately state miss mech. -->

<!-- *Hoe kun je de huidige en nieuwe gereedschappen het beste met elkaar vergelijken* <br> -->

<!-- - data types (e.g. `mids`) -->
<!-- - how 'publication-ready' the visualizations are (e.g. number of lines of code needed* <br>) -->
<!-- - number of functions* <br>  -->
<!-- - number of downloads* <br> `ggmice` has 800 monthly compared to `VIM` 13k and `naniar` 22k -->

<!-- TODO: add complete table with visualization types in all packages to appendix? -->


<!-- |                   Functionaliteit                  |  ggmice  |        VIM       |  naniar  |   amelia  |   mice (lattice)   | -->
<!-- |:--------------------------------------------------|:--------:|:----------------:|:--------:|:---------:|:------------------:| -->
<!-- | Visualisatie van de missingness                    | $\checkmark$       | V                | V        | (beperkt) | V                  | -->
<!-- | Visalisatie van het imputatie EN nonresponse model | V        | X                | X        | X         | X                  | -->
<!-- | Imputatie diagnostiek (visueel) + evt in cijfer?   | V        | V (gedeeltelijk) | X        | V         | V                  | -->
<!-- | Ondersteuning mids en andere data typen            | V        | X                | X        | X         | V (alleen lattice) | -->
<!-- | Grammar of Graphics                                | V        | X                | V        | X         | X                  | -->
<!-- | Aanpasbaarheid                                     | hoog     | laag             | hoog     | laag      | laag               | -->
<!-- | Tidyverse compatible?                              | volledig | beperkt          | volledig | beperkt   | geen               | -->
<!-- | Meteen klaar voor publicatie?                      | hoog     | middel           | hoog     | laag      | laag               | -->


<!-- *Levert het nieuwe gereedschap in de praktijk betere resultaten* <br> -->

<!-- - visualization =/= objective truth -> good imputation model fit/plausible values is not a guarantee for statsistical validity -->
<!-- - visualization is not a replacement of evaluation but support, also check metrics such as FMI -->
<!-- - interpretation remains subjective (e.g. convergence, MAR assumption, etc.) -->

<!-- *Wat vinden toekomstige gebruikers van het gereedschap* <br> -->

<!-- - StackOverflow, GitHub issues, feedback at conferences -->

<!-- *Wat zijn relevante indicatoren voor adoptie* <br> -->

<!-- - CRAN downloads -->
<!-- - GitHub stars and issues -->
<!-- - StackOverflow mentions/questions -->
<!-- - feedback at conferences -->
<!-- - citations of the package (Zenodo reference) -->

<!-- *Welke richtlijnen en advies gelden bij gebruik* <br> -->

<!-- - not a replacement of evaluation but support -->

<!-- *Wat zijn beperkingen van de methode* <br> -->

<!-- - interpretation remains subjective (e.g. convergence, MAR assumption, etc.) -->

<!-- *Wat zijn vragen voor nader onderzoek* <br> -->

<!-- - visualizing `mira` objects -->
<!-- - add posterior predictive check plots & over-imputation -->
<!-- - quantifying uncertainty at pattern/cell level -->
<!-- - tools for sensitivity analyses -->
<!-- - What data-level diagnostics are informative for assessing imputation model adequacy in practice (i.e., when inferential validity cannot be evaluated)? -->
<!-- - Do we get better imputations in practice? -->

